---
title: "GGally_pairs_plots"
author: "Janet Young\n"
date: "`r Sys.Date()`\n"
output: github_document
always_allow_html: true
---

explore [GGally package](https://ggobi.github.io/ggally/articles/ggpairs.html)

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(GGally)

# example data. 244 rows, 7 columns: total_bill  tip    sex smoker day   time size
data(tips)

tips |> summary()
```

Defaults 
- the diagonals give some kind of summary of that variable
- upper and lower panels give different views of the pairwise comparisons:
    - if both variables are continuous, one plot is a scatterplot and the other states the correlation
    - if one continuous and one categorical, we get faceted histograms and boxplots

```{r}
pm <- ggpairs(tips, columns = c("total_bill", "time", "tip"), 
              columnLabels = c("Total Bill", "Time of Day", "Tip"))
pm
```

can group by color
```{r}
pm <- ggpairs(tips, mapping = aes(color = sex), columns = c("total_bill", "time", "tip"))
pm
```




data where log10 would make sense:
```{r}
df_noZeros <- data.frame(x=c(1:50,800,1000), 
                         y=c(1:50,800,1000))
ggpairs(df_noZeros)
```

```{r}
ggpairs(df_noZeros) +
    scale_x_log10() +
    scale_y_log10()
```



data where log10 would make sense, except there are zeros:
```{r}
df_zeros <- data.frame(x=c(0:50,800,1000), 
                         y=c(0:50,800,1000))
ggpairs(df_zeros)
```
add 1 to all columns then plot on a log10 scale:
```{r}
df_zeros |> 
    mutate(across(
        .cols=x:y,
        .fns = function(myNums) { myNums+1 } )
        ) |> 
ggpairs() +
    scale_x_log10() +
    scale_y_log10()
```
```