---
title: "ggplot_demo"
author: "Janet Young\n"
date: "`r Sys.Date()`\n"
output: github_document
always_allow_html: true
---

# Goal

Simply explore the different available geoms


```{r setup, echo=TRUE, results='hide', warning=FALSE, message=FALSE}
## the above is a good chunk header for chunks that load libraries
library(tidyverse)
```


# Example data and base plots

These commands set up the base plots `a` and `b`, but they have no geom layer, so nothing is actually plotted

The `economics` dataset is a 574-row 6-col tibble (wide format) describing how 5 metrics change over time. Column names are `r colnames(economics)`

The `seals` dataset is a 1155-row 4-col tibble that describes how some animals move. Four columns named `r colnames(seals)`


```{r}
# date, pce, pop, psavert, uempmed, unemploy
a <- ggplot(economics, aes(date,unemploy))

# lat  long delta_long delta_lat
b <- ggplot(seals, aes(x=long, y=lat))
# ?seals 
# "This vector field was produced from the data described in..."
### each longitude value in seals is present 21 times, each latitude 55 times
# seals %>% count(long) %>% count(n)
# seals %>% count(lat)  %>% count(n)
# it's simply a grid in x-y space
```


# Explore ggplot geoms

Show some examples from the [ggplot2 cheatsheet](https://rstudio.github.io/cheatsheets/html/data-visualization.html)


geom_blank()

```{r, fig.height=2, fig.width=5}
a + geom_blank() + labs(title="economics, date/unemploy, geom_blank")
```
geom_point()

```{r, fig.height=3, fig.width=5}
b + geom_point(size=0.5) + labs(title="seals, long/lat, geom_point")
```
geom_curve() draws a curved line from (x, y) to (xend, yend). 

aes() arguments: x, xend, y, yend, alpha, angle, color, curvature, linetype, size.

```{r, fig.height=3.5, fig.width=5}
## this is a wierd plot
# x and y already existed in b, and we're adding xend and yend
b + geom_curve(aes(xend = long + 1, 
                   yend = lat + 1), 
               curvature = 1) + 
    labs(title="seals, long/lat, geom_curve")
```
`geom_path()` connects observations in the order they appear.

aes() arguments: x, y, alpha, color, group, linetype, size.


```{r, fig.height=2, fig.width=5}
a + 
    geom_path(lineend = "butt", linejoin = "round", linemitre = 1) +
    labs(title="geom_path")
```

`geom_line()` connects points from left to right. `geom_path()` is similar but connects points in the order they appear in the data. If data are sorted, geom_line() and geom_path plots look the same, otherwise they're different.

```{r, fig.height=2, fig.width=5}
a + 
    geom_path(lineend = "butt", linejoin = "round", linemitre = 1) +
    coord_cartesian(xlim=c(date("1970-01-01"),date("1975-01-01"))) +
    labs(title="geom_path")
# Connect observations in the order they appear. aes() arguments: x, y, alpha, color, group, linetype, size.
```

Same thing with geom_line - looks the same (because data are sorted, geom_path IS connected left-to-right, just like geom_line)

geom_line() - connect observations in the order they appear. a

geom_line aes() arguments: x, y, alpha, color, group, linetype, size.
```{r, fig.height=2, fig.width=5}
a + 
    geom_line(lineend = "butt", linejoin = "round", linemitre = 1) +
    coord_cartesian(xlim=c(date("1970-01-01"),date("1975-01-01"))) +
    labs(title="geom_line")
```

`geom_polygon()` - connect points into polygons

aes() arguments: x, y, alpha, color, fill, group, subgroup, linetype, size.

```{r, fig.height=2, fig.width=5}
a + geom_polygon(aes(alpha = 50)) +
    labs(title="geom_polygon")
```

`geom_rect()` draws a rectangle by connecting four corners (xmin, xmax, ymin, ymax).

aes() arguments: xmax, xmin, ymax, ymin, alpha, color, fill, linetype, size.

the example given makes it hard to see what's going on, so I sample 30 rows at random

```{r, fig.height=2, fig.width=5}
seals %>% 
    slice_sample(n=30) %>% 
    ggplot(aes(xmin = long, ymin = lat, 
               xmax = long + 1, ymax = lat + 1)) + 
    geom_rect() +
    labs(title="geom_rect")
```

`geom_ribbon()` - for each x, plot an interval from ymin to ymax. 

aes() arguments: x, ymax, ymin, alpha, color, fill, group, linetype, size.


```{r, fig.height=2, fig.width=5}
a + geom_ribbon(aes(ymin = unemploy - 900, 
                    ymax = unemploy + 900)) +
    labs(title="geom_ribbon")
```
Plots using `mpg` dataset - a 234 × 11 tibble, showing data for various car types

```{r}
e <- ggplot(mpg, aes(cty,hwy))
```

`geom_point()`

```{r, fig.height=3, fig.width=5}
e + 
    geom_point(size=0.5) +
    labs(title="mpg, geom_point")
```

geom_smooth()

```{r, fig.height=3, fig.width=5}
e + 
    geom_point(size=0.5) + 
    geom_smooth() +
    labs(title="geom_point and geom_smooth")
```



# Finished

```{r}
sessionInfo()
```


