---
title: "ggmsa_useful_tricks_and_testCode"
author: "Janet Young\n"
date: "`r Sys.Date()`\n"
output: github_document
always_allow_html: true
---

[ggmsa](https://www.bioconductor.org/packages/release/bioc/vignettes/ggmsa/inst/doc/ggmsa.html) is a package designed to plot multiple sequence alignments.

ggmsa code does seem to generate a lot of warnings - I am suppressing them here.


# Load libraries

```{r setup, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggtree)
library(ape)
library(Biostrings)
library(ggmsa)
library(here)
```

# Read in some example data

```{r}
## examples adapted from here:
# https://github.com/YuLab-SMU/ggmsa/issues/15
# https://github.com/YuLab-SMU/ggmsa/issues/56
prot_aln_file <- system.file("extdata", "sample.fasta", package = "ggmsa")
prot_aln <- readAAStringSet(prot_aln_file)
dist_matrix <- as.dist(pwalign::stringDist(prot_aln, method = "hamming")/width(prot_aln)[1])
tree <- bionj(dist_matrix)

dna_aln <- readDNAStringSet(here("Rscripts/multiple_sequence_alignments/example_alignment_files/ggmsa_tests_example_for_github.fa"))
```


# Simple ggmsa



```{r, warning=FALSE, message=FALSE}
ggmsa(prot_aln, start=1, end = 50, seq_name=TRUE, border=FALSE) 
```



Wrap a longer alignment over multiple rows (this alignment is ~450aa long)

```{r, fig.height=11, fig.width=7, warning=FALSE, message=FALSE}
ggmsa(prot_aln, seq_name=TRUE, border=FALSE) + 
    facet_msa(field = 100)
```

# Turn background shading off

`none_bg=TRUE` works to turn off the background, but it also turns off the sequence names which is unhelpful

```{r, warning=FALSE, message=FALSE}
ggmsa(prot_aln, start=1, end = 50, none_bg=TRUE, seq_name=TRUE, border=FALSE) 
```

instead we can set up an all-white color scheme and use it in the `custom_color` argument

```{r, warning=FALSE, message=FALSE}
### make a custom color scheme
all_white_df <- data.frame(names = c(LETTERS[1:26],"-"), 
                           color = "#FFFFFF", 
                           stringsAsFactors = FALSE)

ggmsa(prot_aln, start=1, end = 50, custom_color=all_white_df, seq_name=TRUE, border=FALSE) 
```

# Plot tree+aln using ggmsa::geom_msa

```{r, warning=FALSE}
## first make a simple ggtree plot
plain_tree_plot <- ggtree(tree) + 
    geom_tiplab() #tree

## now use ggmsa function tidy_msa() to prep a chunk of the alignment. This returns a data.frame, three columns: seqname, position and amino acid
prot_aln_DF <- tidy_msa(prot_aln, start = 300, end = 330)  

## add alignment to the tree using geom_msa (a ggmsa function)
plain_tree_plot + 
    geom_facet(geom = geom_msa, 
               data = prot_aln_DF,
               panel = 'msa', 
               ## font = NULL, use NULL if we're zoomed way out and only want to show colors
               color = "Chemistry_AA")  +
    xlim_tree(1)
```

# Plot tree+aln using ggtree::msaplot

it's pretty, but I wish it would write the letters of the seqs, rather than just using colors

```{r, warning=FALSE}
msaplot(plain_tree_plot,
        prot_aln_file,
        offset=2, width=3, window=c(200,230))
```


# Add logo plot using geom_seqlogo

```{r, warning=FALSE}
# example from here: https://github.com/YuLab-SMU/ggmsa/issues/56
ggmsa(dna_aln, seq_name = F,
      char_width = 0.5,
      font = NULL, border = NA, color = "Chemistry_NT") +
    geom_seqlogo(color = "Chemistry_NT")
```

# Playing around with the sizing of the plot

I had some trouble with resizing to change dimensions when saving files - played around until I found a solution

```{r, fig.height=3, fig.width=11, warning=FALSE}
test_msa_plot <- ggmsa(dna_aln, seq_name = F,
                       char_width = 0.5,
                       font = NULL, border = NA, color = "Chemistry_NT") +
    facet_msa(field=80) +
    geom_seqlogo(color = "Chemistry_NT")
ggsave(filename = here("Rscripts/multiple_sequence_alignments/ggmsa_testPlot.pdf"), plot=test_msa_plot, height=3, width=11)
test_msa_plot
```

# Try position_highlight

```{r, fig.height=1.2, fig.width=7, warning=FALSE, message=FALSE}
prot_aln[1:2] |> 
    ggmsa(start=1, end=50, 
          position_highlight = c(10, 12),
          seq_name = TRUE)
```

Can have color brightness reflect conservation, although it seems to work with some color schemes and not others. Not sure how useful this is.

```{r}
prot_aln |> 
    ggmsa(start=200, end=250, seq_name = TRUE,
          color = "Hydrophobicity", 
          by_conservation=TRUE)
```


# Maybe we want to add our own layers to the ggmsa


```{r, fig.height=1.2, fig.width=7, warning=FALSE, message=FALSE}
annot_tbl <- tibble(x=c(10,20)) 

prot_aln[1:2] |> 
    ggmsa(start=1, end=50, 
          custom_color=all_white_df, border=FALSE,
          position_highlight = c(),
          seq_name = TRUE) +
    ### tick marks above the alignment
    geom_segment(data=annot_tbl,
                 aes(x=x, xend=x, y=3, yend=4),
                 linewidth=4) +
    ### red highlight
    geom_rect(data=annot_tbl,
              aes(xmin=x-0.5, xmax=x+0.5, ymin=0, ymax=3),
              fill = "red", alpha = 0.3) 
```

# explore the ggplot layers a bit more

```{r}
library(gginnards)
```

```{r, fig.height=2, fig.width=7, warning=FALSE, message=FALSE}
p <- prot_aln[1:5] |> 
    ggmsa(start=200, end=250, seq_name = TRUE)
p
```



That plot has two layers
- layer 1 is the colored tiles
- layer 2 has the letters and x and y axis labels
- x range in each tbl is roughly 1-50 in both
- y range in each tbl is 1-2 in layer 1, 0.7-2.3 in layer 2

[gginnards help](https://cran.rstudio.com/web/packages/gginnards/vignettes/user-guide-2.html)


```{r}
# summary(p)
summary(p$layers)
```


Delete layer 1 (the background) shows us that layer 2 is the letters, x and y axis labels

```{r, fig.height=2, fig.width=7, warning=FALSE, message=FALSE}
delete_layers(p, idx=1L)
```


Delete layer 1 (the colored tiles background) shows us that layer 2 is the letters, x and y axis labels
Deleting layer 2 (the letters etc) shows us that layer 1 is the colored tiles

```{r, fig.height=2, fig.width=7, warning=FALSE, message=FALSE}
delete_layers(p, idx=2L)
```



```{r, warning=FALSE}
## looks like there are 2 tbls
plot_tbls <- lapply(1:2, function(x) {
  layer_data(p, i=x)
})
# get x range and y range
# sapply(plot_tbls, function(one_tbl) {
#   range(one_tbl$x)
# })
# 
# sapply(plot_tbls, function(one_tbl) {
#   range(one_tbl$y)
# })

# sapply(plot_tbls, dim)
#       [,1]  [,2]
# [1,] 13028 13028
# [2,]    15     9
```

First layer tbl (the tiles), first two rows:

```{r}
head(plot_tbls[[1]], 2)
```

Second tbl (the letters, and axes), first two rows:

```{r}
head(plot_tbls[[2]], 2)
```

PANEL is always 1 in each tbl

```{r}
sapply(plot_tbls, function(one_tbl) {
  table(one_tbl$PANEL==1)
})
```



# Finished

```{r}
sessionInfo()
```

