---
title: "test_reroot2"
author: "Janet Young"
date: "`r Sys.Date()`"
output: github_document
always_allow_html: true
---
```{r setup, include=FALSE}
library(ape)
library(ggtree)
library(tidyverse)
```

this is code where I try to understand whether the re-rooting is done CORRECTLY

get an unrooted tree:
```{r}
data(woodmouse)
woodmouse_tree <- nj(dist.dna(woodmouse))
   # 15 tips, 13 internal nodes
```


woodmouse_tree has `r Ntip(woodmouse_tree)` tips, and `r Nnode(woodmouse_tree)` internal nodes

the 'basal' node is node 16 and it has 4 descendents

```{r}
woodmouse_tree |> as_tibble() |> count(parent)
```


```{r}
woodmouse_tree |> as_tibble() |> filter(parent==16)
```

Add fake node labels

```{r}
woodmouse_tree$node.label <- paste("node_", 1:Nnode(woodmouse_tree), sep="")
```


```{r look at it ggtree}
options(ignore.negative.edge=TRUE)
woodmouse_tree |> 
    ggtree() +
    geom_tiplab() +
    geom_nodelab() +
    labs(title="original NJ tree")
```

## reroot woodmouse_tree

this is the bit where I think `root(treedata)` would have screwed up the node labels before. `root(phylo)` is fine though

```{r}
newRoot <- MRCA(woodmouse_tree, c("No1007S","No1103S"))
# numericID 23. I can see from the plot above it's node_8

woodmouse_tree_reroot <- root(woodmouse_tree, 
     node=MRCA(woodmouse_tree, c("No1007S","No1103S")) )
 
options(ignore.negative.edge=TRUE)
woodmouse_tree_reroot |> 
    ggtree() +
    geom_tiplab() +
    geom_nodelab() +
    labs(title="rerooted NJ tree")

```


## with MY data



