---
title: "ggtree_demo"
author: "Janet Young\n"
date: "`r Sys.Date()`\n"
output: github_document
always_allow_html: true
---

```{r setup, include=FALSE}
library(tidyverse)
library(ggtree)
library(treeio)
library(tidytree)
```

# Useful links, general advice

The [ggtree manual](http://yulab-smu.top/treedata-book/index.html) is great (very detailed).
    - Chapter 1 looks at reading/writing trees.  The treeio and ape packages both have tree read/write functions).


# Read a tree and plot it

read in a newick tree file: 
```{r}
nwk_file <- system.file("extdata/sample.nwk", package="treeio") 
nwk_tree <- read.tree( nwk_file )
class(nwk_tree)
nwk_tree
```

plot it using ape's basic plot function
```{r}
plot(nwk_tree)
```

plot it using ggtree
```{r}
nwk_tree %>% 
    ggtree() +
    geom_tiplab()
```

# Associate a tree with additional data

Make a fake data tibble with info on each of the taxa in nwk_tree

```{r}
num_taxa <- length(nwk_tree$tip.label)
tip_dat <- tibble( taxon=nwk_tree$tip.label,
                   fake_height= rnorm(n=num_taxa, mean=100, sd=30 ),
                   fake_phenotype= sample( c("teeth","nose","tail"), 
                                           size=num_taxa, replace=TRUE ),
                   fake_phenotype2=sample( c("fur","scales","spikes"), 
                                           size=num_taxa, replace=TRUE ) )
tip_dat
```

## combine tree and tbl

We associate info tbl with the tree using `left_join` - this creates a single `treedata` object that contains info AND tree (we access the tree like this `nwk_tree_with_info@phylo`) 

Advice:    
- this joining process can be flaky - there's not as much error-checking built-in as I'd like.    
- BEFORE you combine tree with info tbl:   
    - you should manipulate tree to get as close to the final product as you can (e.g. do any rerooting, subsetting, clade rotation, etc, BEFORE you left_join)      
    - manipulate tbl get as close to the final product as you can (e.g. join a bunch of info together)      
    - CHECK taxon labels in tree and the tbl column you plan to join on match up OK    
    - CHECK the taxon ID in the join column of tbl is UNIQUE - repeated values will cause trouble    


```{r}
nwk_tree_with_info <- left_join(nwk_tree, tip_dat, by=c("label"="taxon"))
class(nwk_tree_with_info)
nwk_tree_with_info
```


## plot tree with annotations

now we can use ggtree to use column data in the joined treedata object for labels, colors, etc, etc

```{r}
nwk_tree_with_info %>% 
    ggtree(aes(color=fake_phenotype)) +  # color here colors the branches
    geom_tiplab(aes(color=fake_phenotype)) + # color here colors the tip labels
    geom_tippoint(aes(subset= fake_phenotype=="tail"), color="black") # add dots to taxa, using subset
```


# add heatmap to the right side of a tree

gheatmap is the function for this

## explore gheatmap demo code from ggtree book

gheatmap demo code from the [ggtree online book](http://yulab-smu.top/treedata-book/chapter7.html#gheatmap):

first read in their example tree, `beast_tree`: 

```{r}
beast_file <- system.file("examples/MCC_FluA_H3.tree", package="ggtree")
beast_tree <- read.beast(beast_file)
```

`beast_tree` is a treedata/tidytree object with `r length(beast_tree@phylo$tip.label)` tips


next read in their example genotype data: 

```{r}
genotype_file <- system.file("examples/Genotype.txt", package="ggtree")
genotype <- read.table(genotype_file, sep="\t", stringsAsFactor=F)
colnames(genotype) <- sub("\\.$", "", colnames(genotype))
```

`genotype` is a data.frame with `r dim(genotype)[1]` rows and `r dim(genotype)[2]` columns

`?gheatmap` help page says that the heatmap data must be a matrix or data.frame (not a tibble). I checked - the rownames of `genotype` are identical to the tip labels of the tree, although they're not in the same order


```{r}
# identical(rownames(genotype), beast_tree@phylo$tip.label)
# FALSE
# identical(sort(rownames(genotype)), sort(beast_tree@phylo$tip.label))
# TRUE
```

Now plot:

```{r}
## save a ggtree plot object
p <- ggtree(beast_tree, mrsd="2013-01-01") + 
    geom_treescale(x=2008, y=1, offset=2) + 
    geom_tiplab(size=2)

## add the heatmap to the righthand side. 
gheatmap(p, genotype, offset=15, width=1.5, font.size=3,
         colnames_offset_y= -1) + 
    # scale_y_continuous makes sure we can see the colnames
    scale_y_continuous(expand=c(0.1, 0.1))
```

Same thing but also:    
- add x axis scale bar (and heatmap colnames are less ugly now)    
- use a custom color scheme for the heatmap via `scale_fill_manual`. Default behavior for NAs is that there is no rectangle plotted. If we use `scale_fill_manual` the default behavior for NAs is now different - they are dark gray    

```{r}
p <- ggtree(beast_tree, mrsd="2013-01-01") + 
    geom_tiplab(size=2, align=TRUE, linesize=.5) + 
    theme_tree2()
gheatmap(p, genotype, offset=8, width=0.6, 
         colnames=FALSE, legend_title="genotype") +
    scale_x_ggtree() + 
    scale_y_continuous(expand=c(0, 0.3)) +
    scale_fill_manual(breaks=c("HuH3N2", "pdm", "trig"), 
                      values=c("steelblue", "firebrick", "darkgreen"), 
                      name="genotype")
```

Show that first plot in circle layout (ignore branch lengths)

```{r}
## save a ggtree plot object
p <- ggtree(beast_tree, mrsd="2013-01-01", 
            layout="circular",
            branch.length="none") 

## add the heatmap 
gheatmap(p, genotype, 
         offset=1, 
         width=1, font.size=2,
         colnames_offset_y= -1,
         colnames_angle=270) +
    # scale_y_continuous makes sure we can see the colnames
   scale_y_continuous(expand=c(0.01, 0.01))
```

## gheatmap example with my fake tree and fake data


```{r}
## imagine we're plotting amino acid changes, like Maria's trying to do
# here, NA represents "same as taxon A" (same as human in Maria's case)
# and we'll plot each amino acid as a separate color
tip_dat <- data.frame(row.names=nwk_tree_with_info@phylo$tip.label)
tip_dat$pos1 <- NA
tip_dat$pos2 <- NA
tip_dat$pos3 <- NA
tip_dat[6:8,"pos1"] <- "W"
tip_dat[4:13,"pos2"] <- "T"
tip_dat[9:12,"pos3"] <- "C"

## save a ggtree plot object
p <- nwk_tree_with_info %>% 
    ggtree(aes()) + 
    geom_tiplab(aes())

## add the heatmap to the righthand side. the heatmap data must be matrix or data.frame
# this is the original code
gheatmap(p, tip_dat, 
         offset=3,
         width=0.25, font.size=3) 
```

Or maybe we turn that into WT/ nonWT
```{r}
tip_dat2 <- tip_dat %>% 
    mutate(across(everything(),
                  function(x) {
                      case_when( is.na(x) ~ "WT",
                                 TRUE ~ "nonWT")
                  }))
gheatmap(p, tip_dat2, 
         offset=3,
         width=0.25, font.size=3) +
    scale_fill_manual(breaks=c("WT", "nonWT"), 
                      values=c("lightgray", "firebrick"), 
                      name="")
```


It doesn't work if tip_dat3 is a tibble, although it doesn't give an error, it just doesn't give us a correct heatmap.   Also, setting rownames on a tibble is deprecated.

```{r}
tip_dat3 <- tip_dat2 %>% 
    as_tibble(rownames=NA)
## we can check the rownames are present:
# rownames(tip_dat3)

gheatmap(p, tip_dat3, 
         offset=3,
         width=0.25, font.size=3) +
    scale_fill_manual(breaks=c("WT", "nonWT"), 
                      values=c("lightgray", "firebrick"), 
                      name="")
```

