---
title: "treeTests_miscCodeTroubleshooting"
author: "Janet Young\n"
date: "`r Sys.Date()`\n"
output: github_document
always_allow_html: true
---

```{r setup, include=FALSE}
library(tidyverse)
library(ggtree)
library(treeio)
library(tidytree)
library(ggplot2)
library(TDbook)
```


```{r}
# ### revert back to ggplot 3.5.2 - this is from 2025-04-09 
# ## version 4.0.0 is
# packageurl <- "http://cran.r-project.org/src/contrib/Archive/ggplot2/ggplot2_3.5.2.tar.gz"
# install.packages(packageurl, repos=NULL, type="source")

### revert back to GGally 2.2.1  (2024-02-14) it says it needs 4.4.0
# # GGally_2.2.0.tar.gz
# packageurl <- "http://cran.r-project.org/src/contrib/Archive/GGally/GGally_2.2.1.tar.gz"
# install.packages(packageurl, repos=NULL, type="source")
```


# rotate_tree equal_angle?

Messing around with an equal_angle tree, looking at clipping of taxon labels and rotation

```{r, fig.height=11, fig.width=11}
set.seed(76824375)
tree <- rtree(50)

## make one edge long, to make the tree longer than it is wide
tree$edge.length[1] <- 5

p <- ggtree(tree, layout="equal_angle") + 
    geom_tiplab() +
    labs(title="my_tree")
p
```

ggtree::rotate_tree is not intended for this...

```{r}
rotate_tree(p, 180)
```




I saw a suggestion that in a Rmd doc we can use out.extra='angle=90' to rotate the end result of the plot, but it doesn't seem to work for me:

```{r, out.extra='angle=90'}
p
```
There is a solution using grid newports. Need to mess around more to get it looking right but it's a start


```{r}
library(grid)

### based on 
### https://stackoverflow.com/questions/39712801/how-to-combine-two-ggplots-with-one-rotated

# want to rotate the plot title for this to look good
p2 <- p +
    theme(plot.title = element_text(
        # hjust = 1000, ## hjust seems to have no effect here
        vjust = 1.2,  ## vjust does have an effect though
        angle=270))

grid.newpage()
print(p2, vp=viewport(angle = 90))
```




# Mess around with longer tip labels

```{r}
## make the tip labels longer
tree$tip.label <- paste0("sequence_", tree$tip.label)

```

# other stuff

```{r}
#### tree_anole (phylo object) has 100 tips and 99 internal nodes
### df_svl is a data.frame, 100 rows (named like the tips of tree_anole), 1 column ("svl")
##  `tree_anole` and `df_svl` are from 'TDbook' package

svl <- as.matrix(df_svl)[,1]
fit <- phytools::fastAnc(tree_anole, svl, vars=TRUE, CI=TRUE)

#### make td - data.frame for external nodes
# 100*2 data.frame.
# nodeid() is a tidytree function that converts terminal taxon names to a numeric node ID
td <- data.frame(node = nodeid(tree_anole, names(svl)),
                 trait = svl)
#### make nd - data.frame for internal nodes:
nd <- data.frame(node = names(fit$ace), trait = fit$ace)


#### make d - combined data frame for internal and external nodes
# 199x2 - node IDs and trait value
d <- rbind(td, nd)
d$node <- as.numeric(d$node)

#### add d to the tree to make a treedata object (tree)
## tree is a 'treedata' object - contains the tree (@phylo slot) and a 199*4 tibble generated from d (with an added TRUE/FALSE column called 'isTip')
tree <- full_join(tree_anole, d, by = 'node')

p1 <- ggtree(tree, aes(color=trait), layout = 'circular', 
             ladderize = FALSE, continuous = 'colour', size=2) +
    scale_color_gradientn(colours=c("red", 'orange', 'green', 'cyan', 'blue')) +
    geom_tiplab(hjust = -.1) + 
    xlim(0, 1.2) + 
    theme(legend.position.inside = c(.05, .85)) 

p2 <- ggtree(tree, layout='circular', ladderize = FALSE, size=2.8) + 
    geom_tree(aes(color=trait), continuous = 'colour', size=2) +  
    scale_color_gradientn(colours=c("red", 'orange', 'green', 'cyan', 'blue')) +
    geom_tiplab(aes(color=trait), hjust = -.1) + 
    xlim(0, 1.2) + 
    theme(legend.position.inside = c(.05, .85)) 

plot_list(p1, p2, ncol=2, tag_levels="A")    
```





```{r ggtree example code}


### ggtree manual http://yulab-smu.top/treedata-book/chapter4.html
## treeio https://yulab-smu.top/treedata-book/chapter1.html
## tree
set.seed(2017-02-16)
tree <- rtree(50)

## beast_tree (treedata object, 76 tips, 75 internal nodes)
beast_file <- system.file("examples/MCC_FluA_H3.tree", 
                          package="ggtree")
beast_tree <- read.beast(beast_file)
# ggtree(beast_tree, mrsd="2013-01-01") + theme_tree2()


## NAG (integer vector, length 151, goes with beast_tree)
NAG_file <- system.file("examples/NAG_inHA1.txt", package="ggtree")

NAG.df <- read.table(NAG_file, sep="\t", header=FALSE, 
                     stringsAsFactors = FALSE)
NAG <- NAG.df[,2]
names(NAG) <- NAG.df[,1]

##
nwk <- system.file("extdata/RAxML","RAxML_bipartitions.H3", package='treeio')
tr <- read.tree(nwk)

ggtree(tr) + 
    geom_text2(aes(label=label, 
                   subset = !is.na(as.numeric(label)) & as.numeric(label) > 80))
```

```{r}
raxml_file <- system.file("extdata/RAxML", "RAxML_bipartitionsBranchLabels.H3", package="treeio")
raxml <- read.raxml(raxml_file)
ggtree(raxml) + geom_text(aes(label=bootstrap/100))
```



# Finished

```{r}
sessionInfo()
```

