---
title: "trees_usingDownloadedSpeciesTree"
author: "Janet Young\n"
date: "`r Sys.Date()`\n"
output: github_document
always_allow_html: true
---

```{r setup, include=TRUE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(Biostrings)
library(snakecase)
library(taxize)
library(here)
library(ape)
library(ggtree)
```


I used a mammalian species tree from this paper: Upham, N. S., J. A. Esselstyn, and W. Jetz. 2019. Inferring the mammal tree: species-level sets of phylogenies for questions in ecology, evolution, and conservation. PLOS Biology. https://doi.org/10.1371/journal.pbio.3000494

first I read tree and metadata I downloaded from [this website](https://data.vertlife.org)


metadata file was from [here](https://data.vertlife.org/mammaltree/taxonomy_mamPhy_5911species.csv)

see also notes in `/fh/fast/malik_h/grp/public_databases/trees/Upham_speciesTrees/Upham_speciesTrees_NOTES.txt`


```{r read species tree and metadata}
mammalTreeFiles <- list.files("/Volumes/malik_h/grp/public_databases/trees/Upham_speciesTrees",
                              pattern="nex$", full.names = TRUE)
mammalTrees <- lapply(mammalTreeFiles, ape::read.nexus)
names(mammalTrees) <- gsub("/Volumes/malik_h/grp/public_databases/trees/Upham_speciesTrees/MamPhy_fullPosterior_BDvr_","",mammalTreeFiles)
names(mammalTrees) <- gsub("_MCC_v2_target.nex","",names(mammalTrees))

mammalTreeTaxonomyInfo <- "/Volumes/malik_h/grp/public_databases/trees/Upham_speciesTrees/taxonomy_mamPhy_5911species.csv" %>% 
    read_csv(show_col_types = FALSE)
```

```{r}
## get example species names I want a tree for. This is long-winded but it should be a working example:
RTL3_alnFile <- here("Rscripts/phylogenetics/RTL3_aln2022summer_v28.fa")

RTL3_aln <- readDNAStringSet(RTL3_alnFile)
names(RTL3_aln) <- sapply(strsplit(names(RTL3_aln), " "), "[[", 1)

## now make a seqinfo table - seqname, seqlength, these overlaps, num capsid/protease-overlapping stop-free regions
RTL3_alnInfoTable <- tibble(id=names(RTL3_aln))
## get species name from id
RTL3_alnInfoTable <- RTL3_alnInfoTable %>% 
    separate(id, into=c(NA,"species"), extra="drop", remove=FALSE) %>% 
    mutate(species=str_replace(species, "CapsidHMMhit", "")) %>% 
    mutate(species=str_replace(species, "ProteaseORF", "")) %>% 
    mutate(species=str_replace(species, "BetweenORFsRegion", "")) %>% 
    mutate(species=str_to_sentence(to_snake_case(species))) %>% 
    mutate(species=str_replace_all(species,"_"," "))
```

```{r}
specNamesToGetTreeFor <- RTL3_alnInfoTable %>% 
    select(species) %>% 
    deframe() %>% 
    unique()
```

look up common names for each species and add to RTL3_alnInfoTable

```{r, message=FALSE, warning=FALSE, results=FALSE}
# sci2comm is from the taxize package. it's a bit slow, and very noisy.
commonNameLookup <- sci2comm(sci=specNamesToGetTreeFor)
## I checked - there was 1 character item found for each and none are NA
# class(commonNameLookup)
# table( sapply(commonNameLookup, length))
# table( sapply(commonNameLookup, is.na))

## add commonNameLookup to RTL3_alnInfoTable
commonNameLookup_tbl <- commonNameLookup %>% 
    unlist() %>% 
    as_tibble(rownames="species")
commonNameLookup_tbl <- commonNameLookup_tbl %>% 
    dplyr::rename("common_name"="value")
RTL3_alnInfoTable <- left_join(RTL3_alnInfoTable, commonNameLookup_tbl, by="species") %>% 
    relocate(common_name, .after="species")
```

check they're in the big tree 
```{r}
table(tolower(specNamesToGetTreeFor) %in% gsub("_"," ",tolower(mammalTreeTaxonomyInfo$Species_Name)))
```

they're not all there so I do a bit of wrangling.  google and NCBI taxonomy database helped me find the versions of species names that Upham used
```{r}
RTL3_alnInfoTable <- RTL3_alnInfoTable %>% 
    mutate(upham_name = case_when( species=="Equus asinus" ~ "Equus africanus",
                                   species=="Carlito syrichta" ~ "Tarsius syrichta",
                                   species=="Canis lupus familiaris" ~ "Canis lupus",
                                   species=="Trichechus manatus latirostris" ~ "Trichechus manatus",
                                   species=="Pongo pygmaeus abelii" ~ "Pongo abelii",
                                   species=="Gorilla gorilla gorilla" ~ "Gorilla gorilla",
                                   species=="Cebus imitator" ~ "Cebus capucinus",
                                   species=="Cricetulus griseus" ~ "Cricetulus barabensis",
                                   species=="Piliocolobus tephrosceles" ~ "Colobus guereza", ## not the same species, but I just want it appearing as a sister taxon to  ColobusAngolensism which is present
                                   species=="Nannospalax galili" ~ "Myospalax aspalax", ## really not the same, but any spalax will do
                                   .default = species) )
RTL3_alnInfoTable <- RTL3_alnInfoTable %>% 
    mutate(upham_name=str_replace_all(upham_name, " ", "_"))

## add upham info to my table
RTL3_alnInfoTable <- left_join(RTL3_alnInfoTable, mammalTreeTaxonomyInfo, by=c("upham_name"="Species_Name"))
```

```{r}
### check now whether all the species I want are in the big tree - they are
specNamesToGetTreeFor_upham <- RTL3_alnInfoTable %>%
    # filter(!is.na(name_in_fig)) %>%
    # select(upham_name) %>%
    select(tiplabel) %>% 
    deframe() %>%
    unique()
# table( tolower(specNamesToGetTreeFor_upham) %in% tolower(mammalTreeTaxonomyInfo$tiplabel ))
```

subset the giant tree to get just the species I care about

```{r, fig.height=15, fig.width=7}

## get tree
upham_tree_subset <-  keep.tip(mammalTrees[["DNAonly_4098sp_topoFree_NDexp"]], specNamesToGetTreeFor_upham)

## replace names with common name
upham_tree_subset_rename <- upham_tree_subset
upham_tree_subset_rename$tip.label <- RTL3_alnInfoTable[match(upham_tree_subset$tip.label, RTL3_alnInfoTable$tiplabel),"common_name"] %>% deframe()

### write tree to a file
write.tree(upham_tree_subset_rename, file=here("upham_tree_subset_rename.phy"))

### plot it
upham_tree_subset_rename %>% 
    ggtree() +
    geom_tiplab() +
    hexpand(2)

##  upham_tree_subset_rename_manuallyRotated.nex is where I used figTree to rotate that tree to match species order I wanted and reexported
```

show package versions 
```{r}
sessionInfo()
```


