---
title: "Corruption and human development"
vignette: >
  %\VignetteIndexEntry{Corruption and human development}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r message = FALSE, echo = FALSE}
knitr::opts_chunk$set(
  fig.width = 7,
  fig.height = 4.9
)
```

(his blog: serialmentor.com)

For thisexample, we will use the following packages.
```{r message = FALSE}
library(tidyverse)
library(cowplot)    # for theme_minimal_hgrid()
library(colorspace) # for darken()
library(ggrepel)    # for geom_text_repel() 'everybody should use ggrepel'
```

The dataset is provided as `practicalgg::corruption`. Let's look at it in table form and in basic scatterplot form.
```{r}
corrupt <- practicalgg::corruption %>% 
  filter(year == 2015) %>% 
  na.omit()

corrupt

# aes - how should columns of data map onto aesthetic aspects of plot
# data gets plotted as points
ggplot(corrupt, aes(cpi, hdi, color = region)) +
  geom_point()
```

Basic styling: point colors and theme.

he hates the gray background and gridlines
not a fan of default colors (they are a circle in HCL color space, with same grayness, so look identical if printed grayscale, and they're bad for RG colorblind people)

```{r}
# Okabe Ito colors
region_cols <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#999999")
# the version in the book has 8 colors but we only have 6 here. we have 5 types of region

# there's an aes section that applies to ALL geoms in the plot, and one that applies just to points
ggplot(corrupt, aes(cpi, hdi)) +
  geom_point(
    aes(color = region, fill = region), # has outline AND interior fill.
    # we do NOT specify color by doing e.g. aes(color="green")
    # if we did it OUTSIDE the aes it would work but that's not the point
    # ggplot is set up to use DATA to determine colors
    # we could have done color="Americas" or some other value that is present in the data. ggplot should be clever enough to use the same color for the same data clss in different parts of the plot
    size = 2.5, alpha = 0.5, shape = 21   # 21 is circle, filled.  alpha is transparency
  ) +
  scale_color_manual(
    values = darken(region_cols, 0.3)  # this changes outlines of points. darken is a colorspace function, and we chose our colors here too.  you could also say values=c(Americas="green", etc, etc)
  ) +
  scale_fill_manual(
    values = region_cols    #this is the fill - this is where we choose our colors
  ) +
  theme_minimal_hgrid(12, rel_small = 1) # font size 12 pt throughout 
# theme_minimal_hgrid is a cowplot function.  hgrid = horizontal gridlines. think about which are actually useful. lines likely orthogonal to primary axis of interest. 
# based indirectly on ggplot function called ggtheme (theme_grey)
```

Add smoothing line.

```{r}
## he likes to keep code tidy by putting the geoms near the top and the themes and cosmetic stuff lower
#useful to define explicit scales (e.g. xlims, colors) rather than specify it directly in the code
ggplot(corrupt, aes(cpi, hdi)) +
  geom_smooth(
    aes(color = "y ~ log(x)", fill = "y ~ log(x)"),
    method = 'lm', formula = y~log(x), se = FALSE, fullrange = TRUE
    # se FALSE - no standard err/confidence band.
    # fullrange FALSE - the line would end where data ends
    # on the fly, this created a new column in the data
  ) +
  geom_point(
    aes(color = region, fill = region),
    size = 2.5, alpha = 0.5, shape = 21
  ) +
  scale_color_manual(
    values = darken(region_cols, 0.3)
  ) +
  scale_fill_manual(
    values = region_cols
  ) +
  theme_minimal_hgrid(12, rel_small = 1)
```

Set the same scale name for color and fill scale, to force merging of guides.

```{r}
ggplot(corrupt, aes(cpi, hdi)) +
  geom_smooth(
    aes(color = "y ~ log(x)", fill = "y ~ log(x)"),
    method = 'lm', formula = y~log(x), se = FALSE, fullrange = TRUE
  ) +
  geom_point(
    aes(color = region, fill = region),
    size = 2.5, alpha = 0.5, shape = 21
  ) +
  scale_color_manual(
    name = NULL,  #names NULL so that the two legends get merged
    values = darken(region_cols, 0.3)
  ) +
  scale_fill_manual(
    name = NULL,
    values = region_cols
  ) +
  theme_minimal_hgrid(12, rel_small = 1)
```

Override legend aesthetics.

```{r}
ggplot(corrupt, aes(cpi, hdi)) +
  geom_smooth(
    aes(color = "y ~ log(x)", fill = "y ~ log(x)"),
    method = 'lm', formula = y~log(x), se = FALSE, fullrange = TRUE
  ) +
  geom_point(
    aes(color = region, fill = region),
    size = 2.5, alpha = 0.5, shape = 21
  ) +
  scale_color_manual(
    name = NULL,
    values = darken(region_cols, 0.3)
  ) +
  scale_fill_manual(
    name = NULL,
    values = region_cols
  ) +
  guides(
    color = guide_legend(
      override.aes = list(   #overrides defaults
        linetype = c(rep(0, 5), 1), # to get rid of lines through points in legend
        shape = c(rep(21, 5), NA) # control shapes in legend
      )
    )
  ) +
  theme_minimal_hgrid(12, rel_small = 1)
```

Set x and y scales, move legend on top.
```{r}
ggplot(corrupt, aes(cpi, hdi)) +
  geom_smooth(
    aes(color = "y ~ log(x)", fill = "y ~ log(x)"),  # it calls the y~log(x) thing a sixth column, therefore it uses the sixth color in the color scheme
    method = 'lm', formula = y~log(x), se = FALSE, fullrange = TRUE
  ) +
  geom_point(
    aes(color = region, fill = region),
    size = 2.5, alpha = 0.5, shape = 21
  ) +
  scale_color_manual(
    name = NULL,
    values = darken(region_cols, 0.3)
  ) +
  scale_fill_manual(
    name = NULL,
    values = region_cols
  ) +
  scale_x_continuous(
    name = "Corruption Perceptions Index, 2015 (100 = least corrupt)",
    limits = c(10, 95),  ## these limits will impose limits on the data SO any calculations done during the plotting may NOT include all the data. So, then we would have to do it differently, maybe get the regression outside the plotting code
    breaks = c(20, 40, 60, 80, 100),
    expand = c(0, 0)
  ) +
  scale_y_continuous(
    name = "Human Development Index, 2015\n(1.0 = most developed)",
    limits = c(0.3, 1.05),
    breaks = c(0.2, 0.4, 0.6, 0.8, 1.0),
    expand = c(0, 0)  # to turn off scales expanding outside the limits we set two lines above. 
  ) +
  guides(
    color = guide_legend(
      override.aes = list(
        linetype = c(rep(0, 5), 1),
        shape = c(rep(21, 5), NA)
      )
    )
  ) +
  theme_minimal_hgrid(12, rel_small = 1) +
  theme(
    legend.position = "top",
    legend.justification = "right",
    legend.text = element_text(size = 9),
    legend.box.spacing = unit(0, "pt")
  )
```

Reformat legend into a single row.

```{r}
## mutate is a dplyr function 
corrupt <- corrupt %>%
  mutate(region = case_when(
    region == "Middle East and North Africa" ~ "Middle East\nand North Africa",
    region == "Europe and Central Asia" ~ "Europe and\nCentral Asia",
    region == "Sub Saharan Africa" ~ "Sub-Saharan\nAfrica",
    TRUE ~ region)
    )

ggplot(corrupt, aes(cpi, hdi)) +
  geom_smooth(
    aes(color = "y ~ log(x)", fill = "y ~ log(x)"),
    method = 'lm', formula = y~log(x), se = FALSE, fullrange = TRUE
  ) +
  geom_point(
    aes(color = region, fill = region),
    size = 2.5, alpha = 0.5, shape = 21
  ) +
  scale_color_manual(
    name = NULL,
    values = darken(region_cols, 0.3)
  ) +
  scale_fill_manual(
    name = NULL,
    values = region_cols
  ) +
  scale_x_continuous(
    name = "Corruption Perceptions Index, 2015 (100 = least corrupt)",
    limits = c(10, 95),
    breaks = c(20, 40, 60, 80, 100),
    expand = c(0, 0)
  ) +
  scale_y_continuous(
    name = "Human Development Index, 2015\n(1.0 = most developed)",
    limits = c(0.3, 1.05),
    breaks = c(0.2, 0.4, 0.6, 0.8, 1.0),
    expand = c(0, 0)
  ) +
  guides(
    color = guide_legend(
      nrow = 1,
      override.aes = list(
        linetype = c(rep(0, 5), 1),
        shape = c(rep(21, 5), NA)
      )
    )
  ) +
  theme_minimal_hgrid(12, rel_small = 1) +
  theme(
    legend.position = "top",
    legend.justification = "right",
    legend.text = element_text(size = 9),
    legend.box.spacing = unit(0, "pt")
  )
```

Highlight select countries.

```{r}
country_highlight <- c("Germany", "Norway", "United States", "Greece", "Singapore", "Rwanda", "Russia", "Venezuela", "Sudan", "Iraq", "Ghana", "Niger", "Chad", "Kuwait", "Qatar", "Myanmar", "Nepal", "Chile", "Argentina", "Japan", "China")

corrupt <- corrupt %>%
  mutate(
    label = ifelse(country %in% country_highlight, country, "")  #I think this added a column? or was it there already and it changed it
  )

ggplot(corrupt, aes(cpi, hdi)) +
  geom_smooth(
    aes(color = "y ~ log(x)", fill = "y ~ log(x)"),
    method = 'lm', formula = y~log(x), se = FALSE, fullrange = TRUE
  ) +
  geom_point(
    aes(color = region, fill = region),
    size = 2.5, alpha = 0.5, shape = 21
  ) +
  geom_text_repel(  ## geom_text_repel is a REALLY nice way of labelling points and it avoids overlapping labels by moving text around, and it draws lines to the data points. there are ways to nudge specific labels around (a bit fiddly and unintuitive)
    aes(label = label),
    color = "black",
    size = 9/.pt, # font size 9 pt
    point.padding = 0.1, 
    box.padding = .6,
    min.segment.length = 0,
    seed = 7654
  ) +
  scale_color_manual(
    name = NULL,
    values = darken(region_cols, 0.3)
  ) +
  scale_fill_manual(
    name = NULL,
    values = region_cols
  ) +
  scale_x_continuous(
    name = "Corruption Perceptions Index, 2015 (100 = least corrupt)",
    limits = c(10, 95),
    breaks = c(20, 40, 60, 80, 100),
    expand = c(0, 0)
  ) +
  scale_y_continuous(
    name = "Human Development Index, 2015\n(1.0 = most developed)",
    limits = c(0.3, 1.05),
    breaks = c(0.2, 0.4, 0.6, 0.8, 1.0),
    expand = c(0, 0)
  ) +
  guides(
    color = guide_legend(
      nrow = 1,
      override.aes = list(
        linetype = c(rep(0, 5), 1),
        shape = c(rep(21, 5), NA)
      )
    )
  ) +
  theme_minimal_hgrid(12, rel_small = 1) +
  theme(
    legend.position = "top",
    legend.justification = "right",
    legend.text = element_text(size = 9),
    legend.box.spacing = unit(0, "pt")
  )
```
