---
title: "color_testing"
author: "Janet Young\n"
date: "`r Sys.Date()`\n"
output: github_document
always_allow_html: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# library(tidyverse)
```

Show all color names

```{r}
colors()
```


# use tools from rcolorutils
```{r}
# devtools::install_github("smach/rcolorutils", build_vignettes = TRUE)
library(rcolorutils)
```

rcolorutils provides:     
- the `nearRcolor` function to choose a color by looking in the neighborhood of one you know    
- the `plotCol` function to see those colors    


Show names of colors near deepskyblue - the second arg says what color space we're looking in. The dist is euclidean distance in that color space

```{r}
nearRcolor("deepskyblue", "rgb", dist=50)
```

Display the output of nearRcolor:

```{r, fig.height=1, fig.width=7}
plotCol(nearRcolor("deepskyblue", "rgb", dist=50))
```

Default color space is hsv (with a different dist scale)
```{r, fig.height=1, fig.width=7}
plotCol(nearRcolor("deepskyblue", dist=.1))
```

```{r, fig.height=2, fig.width=7}
plotCol(nearRcolor("tomato", "hsv", dist=.12), nrow=3)
```

I wanted a couple of pinks (for the SATAY analysis)

```{r, fig.height=2, fig.width=7}
plotCol(nearRcolor("pink", dist=0.25), nrow=3)
```

I wanted a couple of light blues (for the SATAY analysis)

```{r, fig.height=2, fig.width=7}
plotCol(nearRcolor("lightblue", dist=0.25), nrow=8)
```


