---
title: "coverage_demo"
author: "Janet Young\n"
date: "`r Sys.Date()`\n"
output: github_document
always_allow_html: true
---



```{r setup, echo=TRUE, results='hide', warning=FALSE, message=FALSE}
## the above is a good chunk header for chunks that load libraries
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)

library(GenomicRanges)
library(GenomicAlignments)

# example data
library(pasillaBamSubset)
```

# Coverage demo

Explore code from [GenomicRanges vignette](https://bioconductor.org/packages/release/bioc/vignettes/GenomicRanges/inst/doc/GenomicRangesHOWTOs.pdf)

We use example data from the pasillaBamSubset - a bam file containing reads mapped to the Drosophila genome.

First, get the bam file name (`un1`), read it in as a `GAlignments` object (`reads1`), and get coverage at each base (`cvg1`) using `GenomicAlignments::coverage()` function.

The result is an RleList object (one item per chromosome). Can get a similar coverage object by importing a bigwig file.

```{r}
un1 <- untreated1_chr4() 
reads1 <- readGAlignments(un1)
cvg1 <- coverage(reads1)
```

This RleList stores coverage for each chromosome, each as a simple Rle object (see `?Rle`):

```{r}
cvg1[["chr4"]]
```

We can get genome info (chromosome lengths) and turn it into a GRanges object representing each chromosome's full length: `genome_gr`

```{r}
genome_gr <- reads1 |> seqinfo() |> GRanges()
genome_gr
```

Use `genome_gr` to get sliding windows for the whole genome (1Mb, sliding by 0.5Mb):

```{r}
genome_1mb_windows <- slidingWindows(genome_gr, width=10^6, step=0.5*10^6) |> 
    unlist()
# The names for each window can be annoying later so I'll strip them off
names(genome_1mb_windows) <- NULL
```


Now we can use the `Views` function to consider coverage for each 1mb window separately

```{r}
Views(cvg1, genome_1mb_windows)
```

And we can feed the result into the `viewMeans()` function to get mean for each window (it returns a list, one per chromosome).

There's also `viewSums()`, `viewMins()`, `viewMaxs()`, and then `viewApply` to do more complex functions for each window.

```{r}
Views(cvg1, genome_1mb_windows) |> 
    viewMeans()
```


I like to store the output (mean coverage) as a column on the original GRanges object - the `unlist` function is useful here:

```{r}
genome_1mb_windows$cov_mean <- Views(cvg1, genome_1mb_windows) |> 
    viewMeans() |> 
    unlist(use.names = FALSE)
```

Show results for just the chr4 windows:

```{r}
genome_1mb_windows[which(seqnames(genome_1mb_windows)=="chr4")]
```


# Finished

```{r}
sessionInfo()
```

