---
title: "explore_purrr"
author: "Janet Young\n"
date: "`r Sys.Date()`\n"
output: github_document
always_allow_html: true
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(patchwork) # for combining plots
```

Learning the `purrr` package

From the [R for Data Science book](https://r4ds.had.co.nz/iteration.html)

`map()` is similar to the `apply()` family of functions.  Flavors of map are defined by what sort of output we want:
    - `map()` makes a list.  
    - `map_lgl()` makes a logical vector.  
    - `map_int()` makes an integer vector.  
    - `map_dbl()` makes a double vector.  
    - `map_chr()` makes a character vector.  

If input is a tbl/data.frame, the `map` functions know to apply a function across each column
```{r}
map_dbl(mtcars, mean)
```

Can use `...` to supply extra arguments


Remember that `.` can be used to refer to the data that's being passed in
```{r}
## this doesn't work (although group_by would)
# mtcars %>% split(cyl)

## this does work, returns list of 3 data frames
# mtcars %>% 
#     split(.$cyl)
```

This could also have been achieved with `group_by %>% summarise`, but for more complex functions you probably need `map`

```{r}
mtcars %>% 
    split(.$cyl) %>% 
    map(function(df){ mean(df$mpg) })
```

```{r}
mtcars %>% 
    group_by(cyl) %>% 
    summarize( mean(mpg) )
```

Using `\` - the backslash is shorthand for `function`:
```{r}
mtcars %>% 
    split(.$cyl) %>% 
    map(\(df){ mean(df$mpg) })
```

A more complex function:  linear modelling using `lm()`. Output of this is a list object, each of which is the output of the lm function:
```{r}
models <- mtcars %>% 
    split(.$cyl) %>% 
    map(function(df) lm(mpg ~ wt, data = df))
```

The `~` ("lambda expression") is now NOT recommended by purrr but you still often see code that uses it online etc.  For example: purrr has a shortcut to make that syntax less verbose - the `.` refers to the current list element, and this is a "one-sided function".  Or, you should use `.x` to refer to the data (and `.y` if there's >1 element)
```{r}
models <- mtcars %>% 
    split(.$cyl) %>% 
    map(~lm(mpg ~ wt, data = .))
```

Extract bits of the output of lm for each list element:
```{r}
models %>% 
    map(summary) %>% 
    map_dbl(~.$r.squared)
```

or this does the same thing:
```{r}
models %>% 
    map(summary) %>% 
    map_dbl("r.squared")
```

selecting list elements by position:
```{r}
x <- list(list(1, 2, 3), list(4, 5, 6), list(7, 8, 9))
x %>% map_dbl(2)
# same as sapply(myList, "[[", 2) 
```

# using map to make a bunch of plots

Here we make a function that generates a plot and test it on the whole dataset

```{r}
## make a function that generates a plot
myPlotFunc <- function(dat) {
    ggplot(dat,
           aes(x = Petal.Width, y = Petal.Length)) +
        geom_point() +
        ggtitle(unique(dat$Species))
}

## test it on the whole dataset
iris %>% myPlotFunc()
```

Now we try using the tidyverse way to split and apply (=map)
```{r}
## group_split then map to plot each (purrr::map)
temp <- iris %>%
    group_split(Species) %>%
    map(myPlotFunc)
temp[[1]] + temp[[2]] + temp[[3]]
```

Another way to write that is by using an anonymous function. Note the `~` before tje function, and the use of `.` within the function to denote the data being passed in

```{r}
temp <- iris %>%
    group_split(Species) %>%
    map(~ggplot(.,
               aes(x = Petal.Width, y = Petal.Length)) +
            geom_point() +
            ggtitle(unique(.$Species)))
temp[[1]] + temp[[2]] + temp[[3]]
```


# Exercises

21.5.3 Exercises (from [here](https://r4ds.had.co.nz/iteration.html))

Write code that uses one of the map functions to:

Compute the mean of every column in mtcars.
```{r}
mtcars %>% 
    map_dbl(mean)
```


Determine the type of each column in nycflights13::flights
```{r}
nycflights13::flights %>% 
    # map(typeof)
    map(class) %>% 
    map_chr(1)  # need this because class(nycflights13::flights$time_hour) had two elements ("POSIXct" "POSIXt")
```


Compute the number of unique values in each column of iris.
```{r}
iris %>% 
    map(unique) %>% 
    map_int(length)
```

Generate 10 random normals from distributions with means of -10, 0, 10, and 100.

```{r}
x <- c(-10, 0, 10, 100)
x %>% 
    map(~ rnorm(n=10, mean=.))
```


How can you create a single vector that for each column in a data frame indicates whether or not it’s a factor?
```{r}
iris %>% 
    map_lgl(is.factor)
```

What happens when you use the map functions on vectors that aren’t lists? What does map(1:5, runif) do? Why?
Answer - it still works, because why not?

What does map(-2:2, rnorm, n = 5) do? Why? 
Answer - it still works.  I actually don't understand how rnorm knows to put the incoming data into the mean argument

What does map_dbl(-2:2, rnorm, n = 5) do? Why?
Answer - it doesn't work, because map_dbl expects only a single number in the output, and rnorm is giving 5 per list element

Rewrite map(x, function(df) lm(mpg ~ wt, data = df)) to eliminate the anonymous function.
```{r}

mtcars %>% 
    split(.$cyl) %>% 
    # map(function(df) lm(mpg ~ wt, data = df)) 
    map(~lm(mpg ~ wt, data = .)) 

```


Dealing with failure

The `safely()` function can be used to wrap any function:

```{r}
safe_log <- safely(log)
safe_log(10)
```

```{r}
safe_log("a")
# or str(safe_log("a"))
```

Here, y is a list of three things, each of which has a result and an error element:
```{r}
x <- list(1, 10, "a")
y <- x %>% map(safely(log))
str(y)
```

purrr's transpose function is nice (but it is old and we should now use list_transpose) - it turns a list-of-lists "inside-out".  now we have a list of two things - all the results, and all the errors:
```{r}
y2 <- y %>% transpose()
str(y2)
```

```{r}
y3 <- y %>% list_transpose()
str(y3)

# can detect errors like this:
is_ok <- y3$error %>% map_lgl(is_null)
# and use that to select input values that caused error:
x[!is_ok]
# or select things that worked: 
# list_c is a bit like unlist - there's also list_flatten list_rbind list_cbind
y3$result[is_ok] %>% list_c()
```


`possibly()` is a bit like `safely()` except that you can define the default output if there's an error:

```{r}
x <- list(1, 10, "a")
x %>% 
    map_dbl(possibly(log, NA_real_))
```

`quietly()` captures other stuff (output, messages, warnings, but not errors - errors will break it)
```{r}
x <- list(1, -1)
x %>% map(quietly(log)) %>% str()
```

Iterating over pairs (or more) of arguments - `map2()` is for pairs of args
```{r}
mu <- list(5, 10, -3)
sigma <- list(1, 5, 10)
map2(mu, sigma, rnorm, n = 5) %>% str()
```

`pmap()` is for groups of >2 args. It's good practise to name the arguments in the list:
```{r}
n <- list(1, 3, 5)
args1 <- list(n=n, mean=mu, sd=sigma)
args1 %>%
    pmap(rnorm) %>% 
    str()
```

args for pmap must always have same length, so they can be stored in a data frame:
```{r}
# tribble() = row-wise tibble creation
params <- tribble(
    ~mean, ~sd, ~n,
    5,     1,  1,
    10,     5,  3,
    -3,    10,  5
)
params %>% 
    pmap(rnorm)
```
`invoke_map()` is used if you want to vary the function that gets used:
```{r}
f <- c("runif", "rnorm", "rpois")
param <- list(
    list(min = -1, max = 1), 
    list(sd = 5), 
    list(lambda = 10)
)
invoke_map(f, param, n = 5) %>% str()
```

that works but invoke_map is deprecated

or using a tibble for the input:
```{r}
sim <- tribble(
    ~f,      ~params,
    "runif", list(min = -1, max = 1),
    "rnorm", list(sd = 5),
    "rpois", list(lambda = 10)
)
sim %>% 
    mutate(sim = invoke_map(f, params, n = 10))
```

except that `invoke_map()` is now deprecated and it tells us to use  map() + exec() instead.  I tried using it but I don't see how to do it.

```{r}
# map(f, exec, !!!param)
# map2(f, list(args1, args2), \(fn, args) exec(fn, !!!args))
# map2(as.list(f), param, exec)
```

use `walk()` when we don't care about the return value, e.g. if we want to render to screen or save files:
```{r}
x <- list(1, "a", 3)

x %>% 
    walk(print)
## the map version of that gives the output of the print function AND returns a list
# x %>% map(print)

```

`walk2` and `pwalk` are more useful

`keep()` and `discard()` are "predicate" functions that help you only do the apply on some columns/list elements:
```{r}
iris %>% 
    keep(is.factor) %>% 
    table()
```

```{r}
iris %>% 
    discard(is.factor) %>%
    map_dbl(mean)
```

`some()` and `every()`:
```{r}
x <- list(1:5, letters, list(10))

x %>% 
    some(is_character)
#> [1] TRUE

x %>% 
    every(is_vector)
#> [1] TRUE

x %>%
    every(is_character)
# FALSE
```


detect() finds the first element where the predicate is true; detect_index() returns its position:
```{r}
x <- sample(10)
x
#  [1]  1  7  4  2  9 10  3  8  5  6

## detect returns the actual value of the first thing that satisfies
x %>% 
    detect(~ . > 5)
#> [1] 7

## detect_index 
x %>% 
    detect_index(~ . > 5)
#> [1] 1

## in this case there are NONE so it fails
x %>% 
    head_while(~ . > 5)

# but this works
x %>% 
    head_while(~ . < 6)

x %>% 
    tail_while(~ . > 4)
```


purrr:::reduce does a function repeatedly (is this like do.call?)
```{r}
dfs <- list(
    age = tibble(name = "John", age = 30),
    sex = tibble(name = c("John", "Mary"), sex = c("M", "F")),
    trt = tibble(name = "Mary", treatment = "A")
)

dfs %>% reduce(full_join)
```
or maybe you have a list of vectors, and want to find the intersection:
```{r}
vs <- list(
    c(1, 3, 5, 6, 10),
    c(1, 2, 3, 7, 8, 10),
    c(1, 2, 3, 4, 8, 9, 10)
)

vs %>% reduce(intersect)
#> [1]  1  3 10
```


reduce() takes a “binary” function (i.e. a function with two primary inputs), and applies it repeatedly to a list until there is only a single element left.

accumulate() is similar but it keeps all the interim results. You could use it to implement a cumulative sum:

```{r}
x <- sample(10)
x
#  [1]  4  5  2  9  6  7  8  1  3 10

## this is the same as cumsum()
x %>% accumulate(`+`)

## this is the same as sum(x):
## x %>% reduce(`+`)  
```

Exercise: Create an enhanced col_summary() that applies a summary function to every numeric column in a data frame.
```{r}

col_summary_2 <- function(df, func) {
    df %>% 
        keep(is.numeric) %>% 
        map(func)
}

iris %>% col_summary_2(mean)

# output <- iris %>% 
#     keep(is.numeric) %>% 
#     map(summary)  %>% 
#     map(~function(dat) { tibble(stat=names(dat), 
#                                 x=as.numeric(dat)) } )
# as_tibble(rownames="stat")
```
