---
title: "mixture_models_learning"
author: "Janet Young\n"
date: "`r Sys.Date()`\n"
output: github_document
always_allow_html: true
---

# Goal

Learn about mixture modelling and explore related R packages.


https://en.wikipedia.org/wiki/Mixture_model

https://yifengedms.github.io/EDMS657-R-Tutorials/Mixture.html

https://mclust-org.github.io/mclust-book/


# General notes

Often we model a mixture of subpopulations that all follow the same type of distribution (e.g. all normal) but it is also possible to model mixtures of different types of distribution.

Expectation maximization algorithms are often used to do the modeling

See also [a script](https://github.com/gak2882/SATAY/blob/main/260615_flowData_analysis/Rscripts_janet/100_flowData_analysis_janet.Rmd) and [its output](https://github.com/gak2882/SATAY/blob/main/260615_flowData_analysis/Rscripts_janet/100_flowData_analysis_janet.Rmd) that I wrote to looks at flow sorting data for the 2micron SATAY project

# Packages used


mclust [vignette](https://cran.r-project.org/web/packages/mclust/vignettes/mclust.html)

mixtools [vignette](https://cran.r-project.org/web/packages/mixtools/vignettes/mixtools.pdf) and [github](https://github.com/dsy109/mixtools) page

[plotmm package](https://packages.oit.ncsu.edu/cran/web/packages/plotmm/vignettes/Getting-Started.html) for plotting model results


```{r setup1, include=FALSE}
## keeping a small first chunk because weirdly Rstudio runs the first chunk at random times and it is slow
knitr::opts_chunk$set(echo = TRUE)
```

```{r setup2, include=FALSE}
library(tidyverse)
library(kableExtra)
library(here) # for tracking project home dir location
library(patchwork)
```


```{r, include=FALSE, message=FALSE}
library(mclust)
library(plotmm)
library(mixtools)
library(datasets)  ## for old faithful data
library(tidyLPA)
library(TidyDensity)
```



# Example data

The classic example dataset is the wait times between eruptions of the Old Faithful geyser, which seems to follow a bimodal distribution. See `?faithful` for more information.

Show the data we are modelling:

```{r, fig.height=3.5, fig.width=8}
wait_histo <- faithful |> 
    ggplot(aes(x=waiting)) +
    geom_histogram(breaks=seq(from=40, to=100, by=5),
                   fill="lightgray", color="black", linewidth=0.2) +
    theme_classic() +
    labs(x="Wait time (minutes)",
         y="number of observations",
         title="(histogram)")

wait_density <- faithful |> 
    ggplot(aes(x=waiting)) +
    geom_density() +
    theme_classic() +
    labs(x="Wait time (minutes)",
         y="Relative frequency",
         title="density plot")

(wait_histo + wait_density) +
    plot_annotation(title="Old Faithful data,\ndistribution of wait time between eruptions")
```


# Try mclust package


Use the `densityMclust` function to model the data. The default setup (for univariate data, like we have) is:

- it returns a `densityMclust` class object
- it models different numbers of classes (components), from 1-9
- it models using equal "E" or unequal ("V") variance
- it can produces a bunch of other plot types, see `?plot.densityMclust`

```{r, fig.height=3.5, fig.width=5, message=FALSE, warning=FALSE}
faithful_mclust <- densityMclust(faithful$waiting,
                                 plot=FALSE, # default is to plot the density of the data
                                 verbose=FALSE)
```

`summary()` is inelegant but shows me the best model:

```{r}
summary(faithful_mclust, parameters = TRUE)
```

We can also ask it to plot the BIC of each model it considered. 

Here we see that two classes is the the most likely solution fit, and that equal variance is slightly more likely than unequal variance.

BIC is a metric that includes some penalty for each additional parameter in the model, so that it tries to avoid overfitting.


We can show the model and the underlying data together:


```{r, fig.height=3.5, fig.width=8, message=FALSE, warning=FALSE}
par(mfrow=c(1,2))
plot(faithful_mclust, what = "BIC")
title(main="model selection", outer=FALSE, line=0.5)

plot(faithful_mclust, what = "density", data = faithful$waiting)
title(main="best model", outer=FALSE, line=0.5)

title(main="mclust modelling", outer=TRUE, line=-2)
```



"Diagnostic" plots show actual and modeled distributions.

```{r, fig.height=3.5, fig.width=8, message=FALSE, warning=FALSE}
par(mfrow=c(1,2))

plot(faithful_mclust, what = "diagnostic", type="cdf")
title(main="CDF diagnostic plot", line=0.5)

plot(faithful_mclust, what = "diagnostic", type="qq")
title(main="QQ diagnostic plot", line=0.5)
```


Made a few attempts to plot the individual components of an mclust model but I don't have it working yet. I bet there's some package that does it that I haven't found yet.

```{r}
## verbose FALSE so we don't get the progress message
faithful_mclust_G2 <- densityMclust(faithful$waiting, G = 2, plot = FALSE, verbose=FALSE)
```

```{r, fig.height=3.5, fig.width=5}
### here's how I can extract the mean and standard deviation of each component and plot them
faithful_x_range_for_density <- extendrange(faithful$waiting, f = 0.3)
faithful_x_values_for_density <- seq(from=faithful_x_range_for_density[1], 
                                     to=faithful_x_range_for_density[2], 
                                     length=1000)

y1 <- dnorm(faithful_x_values_for_density, 
            mean=faithful_mclust_G2$parameters$mean[1], 
            sd=sqrt(faithful_mclust_G2$parameters$variance$sigmasq)) * faithful_mclust_G2$parameters$pro[1]  

y2 <- dnorm(faithful_x_values_for_density, 
            mean=faithful_mclust_G2$parameters$mean[2], 
            sd=sqrt(faithful_mclust_G2$parameters$variance$sigmasq)) * faithful_mclust_G2$parameters$pro[2]  

faithful_mclust_G2_model_to_plot <- tibble(x=faithful_x_values_for_density, 
       y1=y1, 
       y2=y2) |> 
    mutate(y_total=y1+y2)

faithful_mclust_G2_model_to_plot |> 
    ggplot(aes(x=x, y=y1)) +
    geom_histogram(data=faithful, aes(x=waiting, y=after_stat(density)), 
                   fill="lightgray", color="darkgray",
                   binwidth = 5) +
    geom_density(data=faithful, aes(x=waiting, y=after_stat(density)), 
                   color="darkgray") +
    geom_line(color="red") +
    geom_line(aes(x=x, y=y2), color="blue") +
    geom_line(aes(x=x, y=y_total), color="purple") +
    theme_classic() +
    labs(x="Waiting time (minutes)",
         y="Relative frequency",
         title="Old Faithful waiting data and mclust G2 model",
         subtitle=paste0("Gray histogram and density = real data\n",
                         "Red/blue/purple = modeled components"))

```



```{r}
# mclust1Dplot is only for multivariate models
# mclust1Dplot(faithful_mclust_G2)
```

# Try mixtools package 

Do the modelling using `mixtools::normalmixEM`:

- the result (`wait1`) is a `mixEM` object
- `mu` - we provide initial estimates for the means of each distribution. Because we start with a vector of 2 for mu, it models a mix of 2 normal distributions
- `lambda` is the initial mixing proportion. (if you don't specify, it'll start with equal shares)
- it does 9 interations as it optimizes the proportions and means and sigma
- `sigma` is the starting standard deviation
- you can use `mean.constr` to constrain one or more of the means, which could be useful in our case, where we could use the cir0 distribution to guess at one of the components. Same for `sd.constr`

```{r, include=FALSE}
## include=FALSE just so that it doesn't tell me how many iterations it did.
wait1 <- normalmixEM(faithful$waiting, 
                     lambda = .5, 
                     mu = c(55, 80), 
                     sigma = 5)
```

Let's tell it there are 2 categories, but nothing else (normalmixEM does not allow us to try different k all in one call, like mclust does). By default, the variance to be different on the two components but you can constrain it to be the same. You can specify starting mu (mean) and sigma (standard deviation), and you can constrain some components but not others (using mean.constr and sd.constr, see ?normalmixEM)

```{r, include=FALSE}
wait2 <- normalmixEM(faithful$waiting, k = 2)
```



Show a summary of the model:

```{r}
summary(wait1)
```


Plot the model and the input data:

- plot() calls plot.mixEM  (`?plot.mixEM`)
- by default this makes two plots (density and loglik) and requires the user to hit return between each, but we can control that:
- density plot (left) shows the component distributions
- `loglik` plot (right) shows how the log-likelihood of the model changed as the ME iterated - after 2 iterations it didn't improve much

```{r, fig.height=3.5, fig.width=9}
par(mfrow=c(1,2))
plot(wait1, 
     loglik=FALSE, density=TRUE,
     main2="Model (curves) and data (histogram)" ,
     xlab2="Minutes")

plot(wait1, 
     loglik=TRUE, density=FALSE, main1="Log-likelihoods")
```

plotmm::plot_mm can also plot the model (uses ggplot approach). the gray shape is the distribution of the data being modelled, and the colors are the actual data

Works on mixEM objects (e.g. output of mixtools::normalmixEM). Also supports objects modeled by 'EMCluster', and 'flexmix' packages (but not mclust).


```{r, fig.height=3.5, fig.width=5}
p1 <- plot_mm(wait1, 2) +
    labs(title = "normalmixEM model for faithful$waiting",
         subtitle = "Plotted using plot_mm") +
    coord_cartesian(xlim=c(40,100)) 

##  I checked - the gray shows density of the data that was modeled:
# p1 +
#   geom_density(data=faithful, aes(x=waiting), 
#                color="forestgreen", lty=2)

p1
```



Can also customize that using `plot_mix_comps_normal()` as follows:


```{r, fig.height=3.5, fig.width=5}
data.frame(x = wait2$x) |>
    ggplot() +
    ## show distribution of the actual data
    geom_histogram(aes(x, after_stat(density)), 
                   binwidth = 5, colour = "black", fill="lightgray") +
    ### plot component 1
    stat_function(geom = "line", 
                  fun = plot_mix_comps_normal, # here is the function
                  args = list(wait2$mu[1], wait2$sigma[1], lam = wait2$lambda[1]),
                  colour = "red", lwd = 1.5) +
    ### plot component 2
    stat_function(geom = "line", 
                  fun = plot_mix_comps_normal, # here again as k = 2
                  args = list(wait2$mu[2], wait2$sigma[2], lam = wait2$lambda[2]),
                  colour = "blue", lwd = 1.5) +
    ylab("Density") +
    theme_classic()
```


`plot_cut_point()` - if we want to use the model to predict which class each datapoint is in, we might want to determine the thresholds between classes:

```{r, fig.height=3.5, fig.width=5, message=FALSE}
## message=FALSE otherwise I get a message about the binwidth
plot_cut_point(wait1, plot = TRUE, color = "amerika") + # produces plot
    labs(x="Wait time (minutes)")
```

mixtools has a function called multmixmodel.sel that assesses how many components are in the data, but that's only for multivariate data. I don't see an equivalent function for univariate data, unless we have >1 sample of the data. I think maybe we are supposed to subsample for a bootstrapping approach, see ?boot.se



### mixtools::mixturegram

See ?mixturegram

We generate some example data that's a 30:70 mix of two normal distributions, with means of -6 and 0, and standard deviation of 1 for both

```{r}
set.seed(100)
n <- 100
w <- rmultinom(n,1,c(.3,.7))
y <- sapply(1:n, function(i)  {
    w[1,i]*rnorm(1,-6,1) + w[2,i]*rnorm(1,0,1)
})
```

Show the distribution of the data:

```{r, fig.height=3.5, fig.width=5}
tibble(y=y) |> 
    ggplot(aes(x=y)) +
    geom_density() +
    theme_classic()
```

Show a 'mixturegram'. I don't understand what the "PC score" is here.

```{r}
selection <- function(i, data, rep=30){
    out <- replicate(rep,normalmixEM(data,epsilon=1e-06,
                                     k=i,maxit=5000),simplify=FALSE)
    counts <- lapply(1:rep,function(j) 
        table(apply(out[[j]]$posterior,1,
                    which.max)))
    counts.length <- sapply(counts, length)
    counts.min <- sapply(counts, min)
    counts.test <- (counts.length != i)|(counts.min < 5)
    if(sum(counts.test) > 0 & sum(counts.test) < rep) 
        out <- out[!counts.test]
    l <- unlist(lapply(out, function(x) x$loglik))
    tmp <- out[[which.max(l)]]
}

### does mixture modelling with 2:5 classes, replicating each 30 times
all.out <- lapply(2:5, selection, data = y, rep = 2)

## pmbs is a list of length 4, each is a matrix, 100 rows by 2,3,4,5 columns
pmbs <- lapply(1:length(all.out), function(i) 
    all.out[[i]]$post)


mixturegram(y, pmbs, method = "pca", all.n = FALSE,
            id.con = NULL, score = 1, 
            main = "Mixturegram (Well-Separated Data)")

```


# Try tidyLPA



I'm trying the `estimate_profiles` function from the `tidyLPA` package, which is a wrapper for mclust's mixture models

See [tutorial](https://yifengedms.github.io/EDMS657-R-Tutorials/Mixture.html#Example_1:_Univariate_Mixtures) for a very brief outline.

See [tidyLPA intro](https://cran.r-project.org/web/packages/tidyLPA/vignettes/Introduction_to_tidyLPA.html)

```{r}
faithful_tidyLPA <- estimate_profiles(faithful$waiting, 
                              ## this says, try it with 1-6 mixture components
                              n_profiles=1:6,
                              variances = c("equal", "varying"),
                              covariances = c("zero", "zero"))
## by default it says it does the modelling using 'OpenMx' package but it can also use mclust or MplusAutomation. ?estimate_profiles is a bit ambiguous about what the default is, actually.

```

Here's the overall result:

```{r}
faithful_tidyLPA
```


```{r}
faithful_tidyLPA |> 
    compare_solutions()
```

`get_data()` gets the class assignments for each datapoint for each model. Here I choose the best model, and the output has 544 rows (two per observation in faithful$waiting - probability of being in each class)

```{r}
faithful_tidyLPA |> 
    get_data() |> 
    filter(model_number==1 & classes_number==2) |> 
    head()
```


```{r}
p1 <- faithful_tidyLPA |> 
    get_fit() |> 
    mutate(Model=as.factor(Model)) |> 
    ggplot(aes(x=Classes, y=(0-BIC), #group=Model, 
               color=Model)) +
    geom_line() +
    geom_point() +
    theme_classic()+
    labs(title="BIC")
p2 <- faithful_tidyLPA |> 
    get_fit() |> 
    mutate(Model=as.factor(Model)) |> 
    ggplot(aes(x=Classes, y=(0-AIC), #group=Model, 
               color=Model)) +
    geom_line() +
    geom_point() +
    theme_classic() +
    labs(title="AIC")
(p1 + p2) +
    plot_layout(guides="collect")
```

```{r}
faithful_tidyLPA |> 
    get_estimates() |> 
    filter(Model==1 & Classes==2) |> 
    select(-Parameter, -Model, -Classes) |> 
    relocate(Class)
```


```{r}
# faithful_mclust_G2 |> 
     # filter(Model==1 & Classes==2) |> 
    # plot_profiles()
    # plot_density()
```


# TidyDensity package  

`tidy_normal()` might be helpful. 

See [blog post](https://www.spsanderson.com/steveondata/posts/2025-06-23/) or [vignette](https://cran.r-project.org/web/packages/TidyDensity/vignettes/getting-started.html)

```{r, fig.height=3.5, fig.width=5}

tidy_normal(.n = 200, 
            .mean = 0, .sd = 1, 
            .num_sims = 1, .return_tibble = TRUE) |> 
    ggplot(aes(x=y)) +
    geom_density() +
    theme_classic()
```

Use tidy_normal to get a mixture distribution:

```{r, fig.height=3.5, fig.width=5}

dat <- bind_rows(
    tidy_normal(.n = 100, 
                   .mean = 0, .sd = 1, 
                   .num_sims = 1, .return_tibble = TRUE) |> 
        mutate(my_sim_num="A"),
    tidy_normal(.n = 500, 
                   .mean = 10, .sd = 1, 
                   .num_sims = 1, .return_tibble = TRUE)  |> 
        mutate(my_sim_num="B"))

## geom_density default is to scaling so that each group has the same area, 
# use after_stat(count) to scale accordingly
dat |>
    ggplot(aes(x=y, y=after_stat(count), group=my_sim_num, color=my_sim_num)) +
    geom_density(position="stack") +
    theme_classic()
```

# Finished


```{r}
sessionInfo()
```

