---
title: "mixture_models_learning"
author: "Janet Young\n"
date: "`r Sys.Date()`\n"
output: github_document
always_allow_html: true
---

# Goal

Learn about mixture modelling and explore related R packages.


https://en.wikipedia.org/wiki/Mixture_model

https://yifengedms.github.io/EDMS657-R-Tutorials/Mixture.html

## General notes

Often we model a mixture of subpopulations that all follow the same type of distribution (e.g. all normal) but it is also possible to model mixtures of different types of distribution.

Expectation maximization algorithms are often used to do the modeling

```{r setup1, include=FALSE}
## keeping a small first chunk because weirdly Rstudio runs the first chunk at random times and it is slow
knitr::opts_chunk$set(echo = TRUE)
```

```{r setup2, include=FALSE}
library(tidyverse)
library(kableExtra)
library(here) # for tracking project home dir location
library(patchwork)
```


```{r, include=FALSE, message=FALSE}
library(mclust)
library(plotmm)
library(mixtools)
library(datasets)  ## for old faithful data
```



## try mclust 


### mclust on example data



https://cran.r-project.org/web/packages/mclust/vignettes/mclust.html



```{r}
### multivariate
## example data for mclust package
# 145 rows, 4 columns
data(diabetes)

# diabetes |> dplyr::count(class)
#      class  n
# 1 Chemical 36
# 2   Normal 76
# 3    Overt 33
```

```{r}
### univariate

## 155 numeric values
data(acidity)
plot(density(acidity))
```
```{r}
mod4 <- densityMclust(acidity)

```

```{r}
summary(mod4)

```

I think E means the components have equal variance and V means they don't

```{r}
plot(mod4, what = "BIC")

```
Try [plotmm package](https://packages.oit.ncsu.edu/cran/web/packages/plotmm/vignettes/Getting-Started.html) for plotting model results

```{r}
```


```{r}

set.seed(576)

## normalmixEM is from mixtools
mixmdl <- normalmixEM(iris$Petal.Length, k = 2)

# mixmdl$x |> length()

# visualize
plot_mm(mixmdl, 2) +
  labs(title = "Univariate Gaussian Mixture Model",
       subtitle = "Mixtools Object")
```

```{r}
plot_cut_point(mixmdl, plot = TRUE, color = "amerika") # produces plot

```
### mixtools on example old faithful data

```{r}
```

Show the data we are modelling

```{r, fig.width=5, fig.height=3.5}
wait_histo <- faithful |> 
  ggplot(aes(x=waiting)) +
  geom_histogram(breaks=seq(from=40, to=100, by=5),
                 fill="lightgray", color="black", linewidth=0.2) +
  theme_classic() +
  labs(x="Wait time (minutes)",
       y="number of observations",
       title="Old faithful data,\ndistribution of wait time between eruptions")
wait_histo
```

Do the modelling using `normalmixEM`:


-the result (`wait1`) is a `mixEM` object
- `mu` - we provide initial estimates for the means of each distribution. Because we start with a vector of 2 for mu, it models a mix of 2 normal distributions
- `lambda` is the initial mixing proportion. (if you don't specify, it'll start with equal shares)
- it does 9 interations as it optimizes the proportions and means and sigma
- `sigma` is the starting standard deviation
- you can use `mean.constr` to constrain one or more of the means, which could be useful in our case, where we could use the cir0 distribution to guess at one of the components. Same for `sd.constr`

```{r}
wait1 <- normalmixEM(faithful$waiting, 
                     lambda = .5, 
                     mu = c(55, 80), 
                     sigma = 5)
```


```{r}
summary(wait1)
```


Show the model:

```{r}
## plot() calls plot.mixEM 
# ?plot.mixEM
## by default this  makes two plots (loglik and density) and requires the user to hit return between each
## but we can control that and only plot one of those at time
plot(wait1, 
     loglik=FALSE, density=TRUE,
     cex.axis=1.4, cex.lab=1.4, cex.main=1.8, 
     main2="Time between Old Faithful eruptions", 
     xlab2="Minutes")
```

Show how the log-likelihood of the model changed as the ME iterated - after 2 iterations it didn't improve much

```{r}
plot(wait1, 
     loglik=TRUE, density=FALSE,
     cex.axis=1.4, cex.lab=1.4, cex.main=1.8)
```

plot_mm can plot the model. the gray shape is the distribution of the data being modelled, and the colors are the actual data

```{r}
p1 <- plot_mm(wait1, 2) +
  labs(title = "Univariate Gaussian Mixture Model",
       subtitle = "Mixtools Object") +
  coord_cartesian(xlim=c(40,100)) 

##  I checked - tthe gray shows density of the data that was modeled:
# p1 +
#   geom_density(data=faithful, aes(x=waiting), 
#                color="forestgreen", lty=2)

p1
```


```{r}
# Fit a univariate mixture model via mixtools. 
# This time around we allow the sigma (standard dev) to vary - we do tell it to expect 2 components but we don't give any other priors

set.seed(576)

mixmdl <- normalmixEM(faithful$waiting, k = 2)
```



# Customize a plot with `plot_mix_comps_normal()`


```{r}
data.frame(x = mixmdl$x) |>
  ggplot() +
  ## show distribution of the actual data
  geom_histogram(aes(x, after_stat(density)), 
                 binwidth = 5, colour = "black", fill="lightgray") +
  ### plot component 1
  stat_function(geom = "line", 
                fun = plot_mix_comps_normal, # here is the function
                args = list(mixmdl$mu[1], mixmdl$sigma[1], lam = mixmdl$lambda[1]),
                colour = "red", lwd = 1.5) +
  ### plot component 2
  stat_function(geom = "line", 
                fun = plot_mix_comps_normal, # here again as k = 2
                args = list(mixmdl$mu[2], mixmdl$sigma[2], lam = mixmdl$lambda[2]),
                colour = "blue", lwd = 1.5) +
  ylab("Density") +
  theme_classic()
```


# Finished


```{r}
sessionInfo()
```

