---
title: "mutate_across"
author: "Janet Young\n"
date: "`r Sys.Date()`\n"
output: github_document
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

Get example data
```{r}
penguins <- palmerpenguins::penguins
```


Use mutate across with a simple built-in function (example from the help)
```{r}
penguins |>
    mutate(
        across(ends_with("mm"), log10,
               .names = '{.col}_log10'
        )
    ) |>
    select(matches("depth") & (ends_with("mm") | ends_with("log10"))) %>% # remove old columns here
    head(3)
```

Use mutate across with a user-defined function (example from the help)

```{r}
## make a function and use it in mutate across:
divide_by_ten <- function(x) x / 10

## do mutate(across())
penguins |>
    mutate(
        across(
            .cols = ends_with("mm"),   ## everything() chooses all columns
            .fns = divide_by_ten,
            ## use .names arg to figure out what output colnames will be. Empty names arg means values in current columns get replaced
            .names = '{str_remove(.col, "_mm")}_cm'
        )
    ) |>
    select(matches("depth") & (ends_with("mm") | ends_with("cm"))) %>% # remove old columns here
    head(3)
```


Same thing but without defining the function
```{r}
penguins |>
    mutate(
        across(
            .cols = ends_with("mm"),   ## everything() chooses all columns
            .fns = function(x) x / 10,
            ## use .names arg to figure out what output colnames will be. Empty names arg means values in current columns get replaced
            .names = '{str_remove(.col, "_mm")}_cm'
        )
    ) |>
    select(matches("depth") & (ends_with("mm") | ends_with("cm"))) %>% # remove old columns here
    head(3)
```

More practise:

```{r}
myFunc <- function(x) {x^2}
penguins %>% 
    mutate(across( ends_with("_mm"),
                   myFunc,
                   .names = '{.col}_sq'
    )) %>% 
    select(union (ends_with("_mm"), ends_with("_sq") )) %>% 
    head(3)
```

We can also use the ~ style way to specify a function on the fly, in which case we use `.x` as a standin for the data:

```{r}
penguins %>% 
    mutate(across( ends_with("_mm"),
                   ~ (.x ^ 2),
                   .names = '{.col}_sq'
    )) %>% 
    select(intersect(matches("length"), union(ends_with("_mm"), ends_with("_sq") ))) %>% 
    head(3)
```

# A bit more on using `select()`

How to select columns based on more than one condition - use `union`/`intersect`, or `|`/`&`
```{r}
## two alternative ways to write the same thing:
# penguins |>
#     select( ends_with("_mm") | ends_with("_sq")) %>% 
#     head(3)
penguins |> 
    select(union (ends_with("_mm"), ends_with("_g") )) %>% 
    head(3)
```


```{r}
## two alternative ways to write the same thing:
# penguins |>
#     select( ends_with("_mm") & matches("_length") )  %>%
#     head(3)
penguins |>
    select(intersect (ends_with("_mm"), matches("_length") )) %>% 
    head(3)
```


How to select columns based on class of data within them, using `where()`:

Notice that you don’t actually call is.numeric in your code (i.e. there are no paretheses after is.numeric). That’s because you partner up the function is.numeric with the where function. This way, the where function just knows the name of its partner (in this case is.numeric ) and can let its partner do the checking whenever where feels like it.

```{r}
penguins |>
    select(
        where(is.numeric)
    ) %>% 
    head(3)
```


# summarize(across())

Use this instead to get column means, medians etc
```{r}
penguins %>% 
    summarise(across(where(is.numeric), 
                     ~ mean(.x, na.rm=TRUE)))
```
