---
title: "openxlsx_demo"
author: "Janet Young\n"
date: "`r Sys.Date()`\n"
output: github_document
always_allow_html: true
---

# NOTES

The openxlsx library is now really old.

Use one of these packages instead: readxl, writexl, openxlsx2

[This note](https://www.rdocumentation.org/packages/openxlsx/versions/4.2.8.1) is from the openxlsx authors: "openxlsx is no longer under active development. The package is maintained, and CRAN warnings will be fixed, but non-critical issues will not be addressed unless accompanied by a pull request. Packages that depend on openxlsx do not need to take any action, but for new developments, users are encouraged to use alternatives like readxl, writexl, or openxlsx2. The first two packages provide support for reading and writing .xlsx files. The latter package is a modern reinterpretation of openxlsx and provides similar functions to modify worksheets. However, it is not a drop-in replacement, so you may want to consult resources like the update vignette."



# Goal

Show how to use the openxlsx library

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(openxlsx)
```

```{r}
# Tim's example (Hutch Slack channel)
wb <- createWorkbook()
addWorksheet(wb, "SheetName")
freezePane(wb, "SheetName", firstRow = TRUE)
writeDataTable(wb, "SheetName", iris, tableStyle = "TableStyleLight9")
saveWorkbook(wb, "Desktop/temp.xlsx", overwrite = TRUE)

# with my addition
myStyle <- createStyle(fontName="Arial", fontSize = 12, textDecoration="bold")
addStyle(wb, "SheetName", myStyle, stack=TRUE, 
         cols=c(2,3), rows=1:dim(iris)[1], gridExpand=TRUE)
saveWorkbook(wb, "Desktop/temp.xlsx", overwrite = TRUE)


# example from https://cran.r-project.org/web/packages/openxlsx/vignettes/formatting.pdf




## data.frame to write
df <- data.frame("Date" = Sys.Date()-0:4,
                 "Logical" = c(TRUE, FALSE, TRUE, TRUE, FALSE),
                 "Currency" = paste("$",-2:2),
                 "Accounting" = -2:2,
                 "hLink" = "https://CRAN.R-project.org/",
                 "Percentage" = seq(-1, 1, length.out=5),
                 "TinyNumber" = runif(5) / 1E9, stringsAsFactors = FALSE)
class(df$Currency) <- "currency"
class(df$Accounting) <- "accounting"
class(df$hLink) <- "hyperlink"
class(df$Percentage) <- "percentage"
class(df$TinyNumber) <- "scientific"


## Formatting can be applied simply through the write functions
## global options can be set to further simplify things
options("openxlsx.borderStyle" = "thin")
options("openxlsx.borderColour" = "#4F81BD")

## create a workbook and add a worksheet
wb <- createWorkbook()
addWorksheet(wb, "writeData auto-formatting")
writeData(wb, 1, df, startRow = 2, startCol = 2) # df goes into the sheet starting at cell B2
saveWorkbook(wb, "Desktop/temp.xlsx", overwrite = TRUE)

writeData(wb, 1, df, startRow = 9, startCol = 2, borders = "surrounding") # df goes into the sheet again 
saveWorkbook(wb, "Desktop/temp.xlsx", overwrite = TRUE)

writeData(wb, 1, df, startRow = 16, startCol = 2, borders = "rows")
writeData(wb, 1, df, startRow = 23, startCol = 2, borders ="columns")
writeData(wb, 1, df, startRow = 30, startCol = 2, borders ="all")
saveWorkbook(wb, "Desktop/temp.xlsx", overwrite = TRUE)


## headerStyles
hs1 <- createStyle(fgFill = "#4F81BD", halign = "CENTER", 
                   textDecoration = "Bold",
                   border = "Bottom", fontColour = "white")
writeData(wb, 1, df, startRow = 16, startCol = 10, 
          headerStyle = hs1,
          borders = "rows", borderStyle = "medium")
saveWorkbook(wb, "Desktop/temp.xlsx", overwrite = TRUE)

## to change the display text for a hyperlink column just write over those cells
writeData(wb, sheet = 1, x = paste("Hyperlink", 1:5), startRow = 17, startCol = 14)
saveWorkbook(wb, "Desktop/temp.xlsx", overwrite = TRUE)

## writing as an Excel Table
addWorksheet(wb, "writeDataTable")
writeDataTable(wb, 2, df, startRow = 2, startCol = 2)
# TableStyleLight9 etc are defined within Excel
writeDataTable(wb, 2, df, startRow = 9, startCol = 2, tableStyle = "TableStyleLight9")
writeDataTable(wb, 2, df, startRow = 16, startCol = 2, tableStyle = "TableStyleLight2")
writeDataTable(wb, 2, df, startRow = 23, startCol = 2, tableStyle = "TableStyleMedium21")
openXL(wb) ## opens a temp version - pops up an Excel window

saveWorkbook(wb, "Desktop/temp2.xlsx", overwrite = TRUE)



```

