---
title: "patchwork_package_demo_and_tips"
author: "Janet Young\n"
date: "`r Sys.Date()`\n"
output: github_document
always_allow_html: true
---

```{r setup, include=FALSE}
# Load packages 
knitr::opts_chunk$set(echo = TRUE)
library("tidyverse")
library("patchwork")
```

First we make some fake data and plots

```{r make plot objects}
dat <- tibble(x=1:100, y=1:100, type=rep(c("A","B","C","D"), each=25))
p1 <- dat %>% 
    ggplot(aes(x=x, y=y)) +
    geom_point() + 
    labs(title="p1")
p2 <- p1 + labs(title="p2")
p3 <- p1 + labs(title="p3")
```


Show plots in a specific layout:

- use `/` to put plots above each other
- use `|` to put plots side by side
- use `()` to treat plots as a group

```{r}
(p1 / p2) | p3
```

Change widths - note that we NEED the parentheses around the whole plot layout, otherwise the `plot_layout()` function only applies to `p3`, not the whole layout


```{r}
((p1 / p2) | p3) + plot_layout(widths=c(1,3))
```



We can also combine plots using `+` but it's harder to control the layout precisely (we don't always care)

```{r}
p1 + p2 + p3 + plot_layout(ncol=2)
```


Show `sessionInfo()`

```{r}
sessionInfo()
```


