---
title: "table_display_gt_demo"
author: "Janet Young\n"
date: "`r Sys.Date()`\n"
output: github_document
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(gt)
```

Show without adding as_raw_html()

```{r}
# Define the start and end dates for the data range
start_date <- "2010-06-07"
end_date <- "2010-06-14"

# Create a gt table based on preprocessed `sp500` table data
sp500_gt <- sp500 |>
    dplyr::filter(date >= start_date & date <= end_date) |>
    dplyr::select(-adj_close) |>
    gt() |>
    tab_header(
        title = "S&P 500",
        subtitle = glue::glue("{start_date} to {end_date}")
    ) |>
    fmt_currency() |>
    fmt_date(columns = date, date_style = "wd_m_day_year") |>
    fmt_number(columns = volume, suffixing = TRUE) 
sp500_gt

```

Show including as_raw_html()

```{r}
sp500_gt |>
    ## if I don't include as_raw_html(), when I render and display as a github page I get a lot of extra stuff above the table
    as_raw_html()
```

Show sequence alignment as a table - play around with this as an (alternative to ggmsa):

This is a slice of SARS-CoV2's Orf9b prot sequence aligned to related viruses' Orf9b

```{r}
seq_slice <- c("FQLT",
               "FQLI",
               "FQLT",
               "FRLT",
               "FRLI",
               "FRLT",
               "FRLT",
               "FQLT",
               "FQST",
               "FQST")
seq_slice_tbl <- seq_slice %>% 
    strsplit("") %>% 
    as.data.frame() %>% 
    set_names(nm=paste0("seq", 1:10)) %>%
    t() %>% 
    as.data.frame() %>%
    set_names(nm=paste0("pos", 1:4)) %>%
    as_tibble(rownames="id") 
```


Show as a gt table, without using the as_raw_html() function


```{r}
seq_slice_gt <- seq_slice_tbl %>% 
    ## gt alone makes a decent-looking table
    gt() %>% 
    ## a bunch of formatting things:
    cols_align(align = "center", columns=-id) %>% 
    opt_table_font(font = list(google_font(name = "Courier"))) %>% 
    rm_header() %>% 
    tab_options(column_labels.hidden = TRUE,
                table_body.border.bottom.style = "hidden",
                table_body.hlines.style = "hidden",
                data_row.padding=0,
                data_row.padding.horizontal=2,
                ## row striping not visible until I knit. Options don't seem to work. Might be an Rstudio bug
                row.striping.include_table_body = FALSE,
                row.striping.include_stub=FALSE) %>% 
    ## px is number of pixels
    cols_width(id ~ px(75)) 

seq_slice_gt


```



Now add as_raw_html() function

```{r}
seq_slice_gt |>
    ## if I don't include as_raw_html(), when I render and display as a github page I get a lot of extra stuff above the table
    as_raw_html()
```



Maybe I want to use conditional formatting to color certain cells in red ()

```{r}
seq_slice_gt |>
    tab_style(style = cell_text(color = "red"),
              locations = cells_body(columns = pos1, rows = pos1 == "F") ) |>
    tab_style(style = cell_text(color = "red"),
              locations = cells_body(columns = pos2, rows = pos2 == "Q") ) |>
    tab_style(style = cell_text(color = "red"),
              locations = cells_body(columns = pos3, rows = pos3 == "L") ) |>
    tab_style(style = cell_text(color = "red"),
              locations = cells_body(columns = pos4, rows = pos4 == "T") ) |>
    as_raw_html(inline_css=FALSE)
```


# Finished

```{r}
sessionInfo()
```
