---
title: "Untitled"
output: html_document
date: "2025-06-12"
---

# Goal

Temporary script to test things I'm trying to figure out

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(patchwork)
library(ggplot2movies)
```

# violin and stat_summary

```{r}
p <- ggplot(mtcars, aes(factor(cyl), mpg))
p + geom_violin()
```




```{r}
p + 
    geom_violin(aes(fill = factor(am))) + 
    stat_summary(fun = "mean", colour = "red", 
                 position=position_dodge(width=0.9),
                 size = 2, geom = "point")
```


```{r}
mtcars %>% 
    mutate(cyl_char = paste0("cyl",cyl)) %>% 
    mutate(gear_char = paste0("gear",gear)) %>% 
    ggplot(aes(x=factor(cyl_char), 
               y=mpg#, 
               #group=gear_char, 
               #fill=gear_char
               )) +
    geom_violin(scale = "width")
```


# From stat_summary help page


```{r, fig.height=3,fig.width=3}
d <- ggplot(mtcars, aes(cyl, mpg)) + geom_point()
d + stat_summary(fun.data = "mean_cl_boot", colour = "red", linewidth = 2, size = 1)
```

```{r, fig.height=3,fig.width=3}
# Orientation follows the discrete axis
ggplot(mtcars, aes(mpg, factor(cyl))) +
  geom_point() +
  stat_summary(fun.data = "mean_cl_boot", colour = "red", linewidth = 2, size = 1)
```
You can supply individual functions to summarise the value at each x:

```{r, fig.height=3,fig.width=3}

d + stat_summary(fun = "median", colour = "red", size = 2, geom = "point")
```

```{r, fig.height=3,fig.width=3}
d + stat_summary(fun = "mean", colour = "red", size = 2, geom = "point")
```

```{r, fig.height=3,fig.width=3}
d + 
    aes(colour = factor(vs)) + 
    stat_summary(fun = mean, geom="line")
```

```{r, fig.height=3,fig.width=3}
d + stat_summary(fun = mean, fun.min = min, fun.max = max, colour = "red")
```

```{r, fig.height=3,fig.width=3}
d <- ggplot(diamonds, aes(cut))
d + geom_bar()
```

```{r, fig.height=3,fig.width=3}
d + stat_summary(aes(y = price), fun = "mean", geom = "bar")
```

```{r, fig.height=3,fig.width=3}
# Orientation of stat_summary_bin is ambiguous and must be specified directly
ggplot(diamonds, aes(carat, price)) +
  stat_summary_bin(fun = "mean", geom = "bar", orientation = 'y')
```

```{r, fig.height=3,fig.width=3}

# Don't use ylim to zoom into a summary plot - this throws the
# data away
p <- ggplot(mtcars, aes(cyl, mpg)) +
  stat_summary(fun = "mean", geom = "point")
p
```

```{r, fig.height=3,fig.width=3}
p + ylim(15, 30)
```

```{r, fig.height=3,fig.width=3}
# Instead use coord_cartesian
p + coord_cartesian(ylim = c(15, 30))

```

A set of useful summary functions is provided from the Hmisc package:


```{r, fig.height=3,fig.width=3}
stat_sum_df <- function(fun, geom="crossbar", ...) {
  stat_summary(fun.data = fun, colour = "red", geom = geom, width = 0.2, ...)
}
d <- ggplot(mtcars, aes(cyl, mpg)) + geom_point()
# The crossbar geom needs grouping to be specified when used with
# a continuous x axis.
d
```

```{r, fig.height=7, fig.width=7}
p1 <- d + 
    stat_sum_df("mean_cl_boot", 
                mapping = aes(group = cyl)) + 
    labs(title="mean_cl_boot")
p2 <- d + 
    stat_sum_df("mean_sdl", 
                mapping = aes(group = cyl)) + 
    labs(title="mean_sdl")
p3 <- d + 
    stat_sum_df("mean_sdl", 
                fun.args = list(mult = 1), 
                mapping = aes(group = cyl)) + 
    labs(title="mean_sdl with mult=1")
p4 <- d + 
    stat_sum_df("median_hilow", mapping = aes(group = cyl)) + 
    labs(title="median_hilow")

p1 + p2 + p3 + p4 +
    plot_layout(nrow=2)
```



```{r, fig.height=3,fig.width=3}
# An example with highly skewed distributions:

set.seed(596)
mov <- movies[sample(nrow(movies), 1000), ]
 m2 <-
   ggplot(mov, aes(x = factor(round(rating)), y = votes)) +
   geom_point()
 m2 <-
   m2 +
   stat_summary(
     fun.data = "mean_cl_boot",
     geom = "crossbar",
     colour = "red", width = 0.3
   ) +
   xlab("rating")
m2

```


Notice how the overplotting skews off visual perception of the mean supplementing the raw data with summary statistics is _very_ important

Next, we'll look at votes on a log scale.

Transforming the scale means the data are transformed first, after which statistics are computed:

```{r, fig.height=3,fig.width=3}
m2 + scale_y_log10()
```

Transforming the coordinate system occurs after the statistic has been computed. This means we're calculating the summary on the raw data and stretching the geoms onto the log scale.  Compare the widths of the standard errors.

```{r, fig.height=3,fig.width=3}
m2 + coord_trans(y="log10")
```

```{r}

```

